---
title: "Food Nutrition and Price"
author: Duncan Garmonsway
description: |
  Explore a supermarket of food nutrition and price (renders slowly)
date: January 7, 2019
preview: tweet.png
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---

After feasting, thoughts turn to diets.  Diets mean optimisation.  Optimisation
requires data.  For myself, I want to know:

* What is the cheapest protein, fat and carbohydrate per 100g?
* What is the general distribution of nutritional value?
* What is the sweeter condiment: sauce or jam?

So I scraped a supermarket.

Specifically, I scraped the grocery section of the website of a major UK
supermarket, and then I sat on the data for nearly two years because it's so
untidy.

## Getting nutritional information in bulk

There are strict [labelling requirements on
food](https://www.gov.uk/food-labelling-and-packaging), its nutritional value
in particular, so almost all food packaging gives carbohydrate, fat and protein
content per 100g or 100ml.  This should make it straightforward to learn about
food, because the data already exists.

> ![Table of nutritional information for Sainsbury's Brocolli Cheese ready
meal](./nutritional-information.png)

Unfortunately, the government doesn't collect the data, though the Office for
National Statistics [scraped three supermarked
websites](https://www.ons.gov.uk/file?uri=/economy/inflationandpriceindices/articles/researchindicesusingwebscrapedpricedata/may2016update/researchindicesusingwebscrapeddatamay2016.pdf)
for the Consumer Price Index.  I tried to do the same and regretted that the ONS
hadn't pubished its code, because it was a pain.  A couple of packages have made
the task tractable; [htmltab](https://cran.r-project.org/package=htmltab) is
helps with spanning rows and columns html tables, my own
[unpivotr](https://nacnudus.github.io/unpivotr) has improved since 2016, and
[reticulate](https://rstudio.github.io/reticulate/index.html) makes it trivial
to call the Python package [quantulum3](https://github.com/nielstron/quantulum3)
for extracting units like grams and kilocalories from free text.  The result
isn't perfectly clean data -- you'll notice some obvious glitches -- but it's
good for shopping.

<aside>
This is the least reproducible work I have ever done.  If
I were to do it again I would use the
[`polite`](https://github.com/dmi3kno/polite) to be nice to the supermarket, and
[`furrr`](https://cran.r-project.org/package=furrr) to get it done faster (it
ran overnight).  I haven't published the data because I don't think it belongs
to me, but if you really want it then contact me.  You could get some of it from
this web page anyway, but you won't get the vitamins and minerals etc. in the
raw data.
</aside>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(scales)
library(crosstalk)
library(plotly)
library(DT)
library(here)

nutrients <- readRDS(here("data", "nutrients.Rds"))

prices <-
  readRDS(here("data", "prices.Rds")) %>%
  dplyr::filter(!is.na(price_per_100g)) %>%
  group_by(filename) %>%
  slice(1L) %>%
  ungroup() %>%
  rename(product = filename) %>%
  mutate(product = str_replace(product, "^.*\\|", ""),
         product = str_replace(product, "\\.html$", "")) %>%
  select(product, price_per_100g) %>%
  rename(price = price_per_100g)

basic_nutrients <-
  nutrients %>%
  dplyr::filter(!str_detect(filename, fixed("sainsburys-indian-meal--taste-the-difference-143kg-%28serves-2%29"))) %>%
  dplyr::filter(unit %in% c("gram", "kilocalorie"),
                !(unit == "gram" & amount > 100),
                !(nutrient == "energy" & amount > 1000),
                str_detect(nutrient,
                           regex("carbohydrate|energy|fat|protein|sugar",
                                 ignore_case = TRUE))) %>%
  # left_join(prices_per_100g, by = "filename") %>%
  rename(product = filename) %>%
  mutate(product = str_replace(product, "^.*\\|", ""),
         product = str_replace(product, "\\.html$", ""),
         nutrient = str_extract(nutrient,
                                regex("carbohydrate|energy|fat|protein|sugar", ignore_case = TRUE))) %>%
  select(-unit) %>%
  group_by(product, nutrient) %>%
  slice(1) %>%
  ungroup() %>%
  spread(nutrient, amount) %>%
  replace_na(list(carbohydrate = 0,
                  energy = 0,
                  fat = 0,
                  protein = 0,
                  sugar = 0)) %>%
  dplyr::filter(energy > 0) %>%
  select(product, carbohydrate, fat, protein, energy) %>%
  left_join(prices, by = "product") %>%
  mutate(`£/carbohydrate` = price / carbohydrate,
         `£/fat` = price / fat,
         `£/protein` = price / protein,
         `£/energy` = price / energy)
```

## The cheapest food by nutritional value

The cheapest grocery overall, by mass, was own-brand diet lemonade.  All the top
ten were soft drinks.  But what you're buying when you buy diet pop is basically
water, so it's far from the cheapest for any kind of nutrition.  The cheapest
energy and cheapest carbohydrate is flour, with bread and sugar filling out the
top ten (wine is probably a mistake).  The cheapest fat is lard, with non-butter
spread, oil, and mayonnaise (mostly oil) filling out the top ten.

The cheapest protein?  Also flour.  But you could splash out and get 25kg of
chicken portions.

Expand the headings below to see the top 10 in each category.

<details><summary>Price per unit mass</summary>

```{r price-mass}
basic_nutrients %>%
  arrange(price) %>%
  slice(1:10)
```
</details>

<details><summary>Price per unit energy</summary>

```{r price-energy}
basic_nutrients %>%
  arrange(`£/energy`) %>%
  slice(1:10)
```
</details>

<details><summary>Price per unit carbohydrate</summary>

```{r price-carbohydrate}
basic_nutrients %>%
  arrange(`£/carbohydrate`) %>%
  slice(1:10)
```
</details>

<details><summary>Price per unit fat</summary>

```{r price-fat}
basic_nutrients %>%
  arrange(`£/fat`) %>%
  slice(1:10)
```
</details>

<details><summary>Price per unit protein</summary>

```{r price-protein}
basic_nutrients %>%
  arrange(`£/protein`) %>%
  slice(1:10)
```
</details>

<p/>

Unsurprisingly, protein is the most expensive nutritional component, then fat,
then carbohydrate.  But now you can tell when you're paying over the odds.
There's a lot of choice for less than 10p per gram of protein or fat, or less
than 1p per gram of carbohydrate.  If you're paying more than £1 per gram of
anything then it isn't for the nutritional value (unless it's for something I
haven't explored, such as fibre and vitamins).

```{r price, warning = FALSE}
basic_nutrients %>%
  select(product, starts_with("£"), -`£/energy`) %>%
  gather(unit, price, starts_with("£")) %>%
  mutate(unit = str_sub(unit, 3L)) %>%
  ggplot(aes(price, fill = unit)) +
  geom_density(alpha = .5, colour = NA) +
  scale_x_log10(labels = dollar_format(prefix = "£"),
                limits = c(NA, 100),
                breaks = log_breaks()(c(1e-2, 1e-1, 1, 1e1))) +
  scale_fill_brewer(name = "", palette = "Dark2") +
  xlab("Price per gram (log scale)") +
  ylab("") +
  theme_minimal() +
  ggtitle("Price of food by nutrition",
          subtitle = "Distribution of price per gram of carbohydrate, fat and protein") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = c(0.7, .7))
```

## Distribution of nutritional value

You can explore an interactive 3d graph at the bottom of the post, but first I
want to use ternary plots to compare particular groups of foods.  A little light
curation before taking the lazy option.

Ternary plots compress three-dimensional data onto two dimensions by exploiting
a loophole: when the dimensions are ratios that add to 100%, there are only two
degrees of freedom.  Once you know the proportions of two dimensions, then you
know the third.  This doesn't work for carbohydrate, fat and protein because
they don't necessarily sum to 100% of the mass of the food -- diet soda drinks
are 0% everything.  So to make the food fit the graph, I create fake ratios:
carbohydrate divided by (carbohydrate plus fat plus sugar), and so on.  When
reading the graphs below, a point that is near to the "carbohydrate" corner has
more carbohydrate than fat or protein, but it doesn't necessarily have much of
any of them -- it could be mostly water.  Finally, the colour is the amount of
energy, with turquoise for lots of energy and purple for not much (or missing
data).  Mouse over for the names of the products.

```{r ternary-data, include = FALSE}
ternary_df <-
  basic_nutrients %>%
  select(product, price, energy, carbohydrate, fat, protein) %>%
  # get carb, fat, protein as proportions of carb + fat + protein
  mutate(sum = (carbohydrate + fat + protein) / 100) %>%
  gather(nutrient, amount, carbohydrate, fat, protein) %>%
  group_by(product, energy, sum) %>%
  mutate(amount = amount / sum(amount)) %>%
  ungroup() %>%
  replace_na(list(amount = 0)) %>%
  spread(nutrient, amount)
```

```{r ternary-energy, warning = FALSE}
ternary_df %>%
  plot_ly() %>%
  add_trace(
    type = "scatterternary",
    mode = "markers",
    marker = list(size = 2,
                  color = ~energy,
                  colorscale = "Viridis"),
    a = ~carbohydrate,
    b = ~fat,
    c = ~protein,
    text = ~product
  ) %>%
  layout(
    ternary = list(
      sum = 100,
      aaxis = list(title = "carbohydrate"),
      baxis = list(title = "fat"),
      caxis = list(title = "protein")
    )
  )
```

You might wonder about a data glitch in that plot.  Why do there seem to be so
many foods that despite having proportions of carbohydrate don't have much
energy?  That's because the purple foods on the right are comparitively high in
water content -- meat, bread, soup, cooked rice and pasta -- whereas the
turquoise foods on the left are more substantial -- biscuits, crisps, sticky
cake.  Of course there is probably some missing data too.

```{r ternary-pairplot, include = FALSE}
ternary_pairplot <- function(.data) {
  .data %>%
    plot_ly(color = ~group, colors = c("#F8766D", "#00BFC4")) %>%
    add_trace(
      type = "scatterternary",
      mode = "markers",
      marker = list(size = 4), # opacity per point isn't supported
      a = ~carbohydrate,
      b = ~fat,
      c = ~protein,
      text = ~product
    ) %>%
    layout(
      ternary = list(
        sum = 100,
        aaxis = list(title = "carbohydrate"),
        baxis = list(title = "fat"),
        caxis = list(title = "protein")
      )
    )
}
```

Ternary plots are good for comparing groups of foods, and here they reveal some
surprises.  Crisps and biscuits are roughly equivalent, nutritionally.  Some
crisps are fattier, and some biscuits are more sugary, but there is a
substantial overlap.

```{r crisps-biscuits-plotly}
ternary_df %>%
  mutate(group = str_extract(product, "(crisps)|(biscuits)")) %>%
  dplyr::filter(!is.na(group)) %>%
  ternary_pairplot()
```

You might pay closer attention to food labelling when you compare sauces with
jams.  Fair enough, most of the jams are overplotted in the carbohydrate corner,
but lots of sauces are terribly sugary.

<aside>
The data distinguishes between carbohydrates and "of which, sugar", but I
haven't here.
</aside>

```{r jam-sauce}
basic_nutrients %>%
  mutate(group = str_extract(product, "(-jam-)|(sauce)")) %>%
  dplyr::filter(!is.na(group)) %>%
  ternary_pairplot()
```

'Lite' yoghurt is not so much less fatty as less carbohydrate-y than normal
yoghurt.

```{r yoghurt}
basic_nutrients %>%
  dplyr::filter(str_detect(product, "yog[h]?urt")) %>%
  mutate(group = str_extract(product, "(fat-free)|(low-fat)|(sugar-free)|(low-sugar)|(light)|(lite)")) %>%
  replace_na(list(group = "normal")) %>%
  mutate(group = if_else(group == "normal", "normal", "lite")) %>%
  ternary_pairplot()
```

## Free exploration

It's your turn to explore the data.  Perhaps you'd like to check that my
findings above are still valid as absolutes rather than proportions.  Or perhaps
you've resolved to bulk up and want to sort the data by protein content (egg
white powder, then dried meat).  The highest energy content?  Peppermint extract
is in sixth place, being almost 100% fat.  You can filter the table, e.g.  to
find the lowest-sugar sugar (molasses, 94%), and click on rows (hold down shift
to select a range) to filter the charts as well.  Finally, select a region of
the 2d graphs to filter the other graphs and the table.  Be patient -- the
browser has a lot of data to deal with.

<aside>
The interactivity within graphs [plotly](https://plot.ly/r/) and the
interactivity between graphs and the table is
[crosstalk](https://rstudio.github.io/crosstalk/).  Everything runs in the
browser; there is no server.  [Matt Dray's blog
post](https://www.rostrum.blog/2018/09/12/crosstalk-memes/) and
[slides](https://matt-dray.github.io/earl18-presentation/#1) are persuasive and
fun.
</aside>

```{r 3d, warning = FALSE}
food_df <- SharedData$new(basic_nutrients, ~product, group = "food")

food_df %>%
  plot_ly(x = ~carbohydrate, y = ~fat, z = ~protein, color = ~energy,
          type = "scatter3d",
          mode = "markers",
          marker = list(size = 1.25),
          text = ~product) %>%
  layout(scene = list(xaxis = list(title = 'carbohydrate'),
                      yaxis = list(title = 'fat'),
                      zaxis = list(title = 'protein'),
                      dragmode = "turntable")) %>%
  highlight(opacityDim = 0.01, on = "plotly_selected", off = "plotly_deselect")

```

```{r 2d, fig.width = 9.5}
food_df %>%
  # Control y-axis lengths until https://github.com/plotly/plotly.js/issues/1876
  plot_ly(height = figWidth() / 3) %>%
  add_trace(x = ~carbohydrate, y = ~fat, color = ~energy,
            type = "scatter",
            mode = "markers",
            marker = list(size = 1.25),
            text = ~product) %>%
  add_trace(x = ~protein, y = ~carbohydrate, color = ~energy,
            xaxis = "x2",
            yaxis = "y2",
            type = "scatter",
            mode = "markers",
            marker = list(size = 1.25),
            text = ~product) %>%
  add_trace(x = ~fat, y = ~protein, color = ~energy,
            xaxis = "x3",
            yaxis = "y3",
            type = "scatter",
            mode = "markers",
            marker = list(size = 1.25),
            text = ~product) %>%
  layout(xaxis = list(title = "carbohydrate",
                      domain = c(0, 0.3),
                      range = c(0, 100)),
         xaxis2 = list(title = "protein",
                       domain = c(0.33, 0.66),
                       range = c(0, 100)),
         xaxis3 = list(title = "fat",
                       domain = c(0.7, 1),
                       range = c(0, 100)),
         yaxis = list(title = "fat",
                      domain = 0:1,
                      range = c(0, 100),
                      anchor = "x",
                      scaleanchor = "x"),
         yaxis2 = list(title = "carbohydrate",
                       domain = 0:1,
                       range = c(0, 100),
                       anchor = "x2",
                       scaleanchor = "x2"),
         yaxis3 = list(title = "protein",
                       domain = 0:1,
                       range = c(0, 100),
                       anchor = "x3",
                       scaleanchor = "x3"),
         showlegend = FALSE) %>%
  highlight(opacityDim = 0.01,
            on = "plotly_selected",
            off = "plotly_doubleclick") %>%
  hide_colorbar()
```

```{r datatable, layout = "l-screen-inset"}
datatable(food_df)
```
