---
title: "CRAN h-index"
description: |
  h-indexes of CRAN package maintainers.  Read to the end to find your own
  h-index.
author: Duncan Garmonsway
date: February 15, 2019
output:
  distill::distill_article:
    self_contained: false
preview: cran-h-index_files/figure-html5/graph-1.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE,
                      echo = TRUE)

library(tidyverse)
library(cranly)
library(DT)
library(here)
```

People have blogged before about [the most prolific package maintainers on
CRAN](https://blog.revolutionanalytics.com/2018/03/the-most-prolific-package-maintainers-on-cran.html)
and the [top five R package
maintainers](https://www.datacamp.com/community/blog/top-5-r-package-maintainers).
But _quantity_ of publications is not the same as _quality_.  An alternative
metric called the [*h*-index](https://en.wikipedia.org/wiki/H-index) summarises
both the total number of a person's publications (quantity) as well as the
number of times each publication has been cited in other publications (quality,
kinda).

> The *h*-index is an author-level metric that attempts to measure both the
> productivity and citation impact of the publications of a scientist or
> scholar. The index is based on the set of the scientist's most cited papers
> and the number of citations that they have received in other publications. The
> index can also be applied to the productivity and impact of a scholarly
> journal[1] as well as a group of scientists, such as a department or
> university or country.[2] The index was suggested in 2005 by Jorge E. Hirsch,
> a physicist at UC San Diego, as a tool for determining theoretical physicists'
> relative quality[3] and is sometimes called the Hirsch index or Hirsch number.
> -- [Wikipedia](https://en.wikipedia.org/wiki/H-index)

What would an H-index for CRAN packages look like?  I used the `DESCRIPTION`
files of CRAN packages for some definitions:

* Authors: people or companies listed in the `Author` and `Maintainer` fields.
* Citations: packages listed in the `Depends` and `Imports` fields.

`DESCRIPTION` files are available as structured data via
`tools::CRAN_package_db()`.  The `Authors` field isn't really structured, but
the [`cranly`](https://github.com/ikosmidis/cranly) package (GitHub only) does
impressive work with RegEx to tidy it up.  Of course it can't do anything about
[`Author: See AUTHORS
file.`](https://github.com/ropensci/git2r/blob/master/DESCRIPTION)

<aside>

'Maintainer' is structured because CRAN uses it in an
[API](https://cran.r-project.org/web/checks/check_summary_by_maintainer.html).
'Author' isn't structured because CRAN doesn't use it in an API.  Lesson: the
best data is the data that is used at the source.

</aside>

## Preparation

```{r DESCRIPTION}
pkglist <-
  tools::CRAN_package_db() %>%
  cranly::clean_CRAN_db() %>%
  as_tibble()
```

<details><summary>Expand to glimpse the data</summary>

```{r DESCRIPTION-glimpse}
glimpse(pkglist)
```

</details>

<p/>

From that data you can combine the names of packages listed in `depends` and
`imports`.

```{r combine-depends-imports}
# Depends and Imports count, Suggests and Enhances doesn't.
packages <-
  pkglist %>%
  mutate(maintainer = clean_up_author(maintainer)) %>%
  select(package, maintainer, author, depends, imports) %>%
  mutate(author = map2(maintainer, author, c),
         cites = map2(depends, imports, c)) %>%
  select(package, author, cites) %>%
  print()
```

That data frame has to be unnested in stages because both `author` and
`cites` are ragged list-columns.  Separate data frames are created, one
for the package--author pairs, the other for package--cited-package pairs.  Then
they are joined to make a data frame with one row per package per author per
citation.

```{r split-authors-citations}
package_citations <-
  packages %>%
  unnest(cites) %>%
  dplyr::filter(!is.na(cites)) %>% # NAs occur when depends or imports is empty.
  distinct() %>% # Duplicates occur when a package is in both depends and imports.
  print()

package_authors <-
  packages %>%
  unnest(author) %>%
  distinct() %>% # Duplicates occur when a maintainer is also an author (often)
  print()
```

Recombining authors and citations.  I did some extra work to detect
self-citations.

```{r recombine-authors-citations}
citations <-
  package_authors %>%
  inner_join(package_citations, by = c("package" = "cites")) %>%
  rename(cited_by_package = package.y) %>%
  left_join(package_authors, by = c("cited_by_package" = "package")) %>%
  rename(author = author.x,
         cited_by_author = author.y) %>%
  mutate(self_citation = cited_by_author == author) %>%
  print()
```

Next this needs to be aggregated to one row per author per package, and the
number of times that package is cited by other packages. Again I complicated
this by doing it twice, once including self-citations and once excluding them,
then joining the two.  This is now ready for h-indexing.

```{r count-citations}
author_citations_inc_self <-
  citations %>%
  distinct(package, author, cited_by_package) %>%
  count(package, author) %>%
  rename(citations_inc_self = n)

author_citations_exc_self <-
  citations %>%
  dplyr::filter(!self_citation) %>%
  distinct(package, author, cited_by_package) %>%
  count(package, author) %>%
  rename(citations_exc_self = n)

author_citations <-
  inner_join(author_citations_inc_self,
             author_citations_exc_self,
             by = c("package", "author")) %>%
  right_join(package_authors, by = c("package", "author")) %>%
  mutate_if(is.numeric, replace_na, 0L) %>%
  print()
```

## *h*-index

With the scutwork done, brace yourself for the fun algorithmic bit, the
insignificant fraction of the code that calculates the h-indexes.  For each
author, it calculates the largest `n` such that the author has `n` packages each
cited at least `n` times.  So if their *h*-index is 10, then they have 10
packages each cited at least 10 times, and they don't have 11 packages each
cited at least 11 times.

```{r define-h-index}
h_index <- function(citations_per_package) {
  packages_per_n_citations <- table(citations_per_package)
  n_citations <- c(0L, seq_len(max(citations_per_package)))
  n_packages <- integer(length(n_citations))
  n_packages[as.integer(names(packages_per_n_citations)) + 1] <-
    packages_per_n_citations
  n_gte_n <- rev(cumsum(rev(n_packages))) # no. pkgs with >= no. citations
  tail(c(0L, n_citations[n_gte_n >= n_citations]), 1L)
}
```

Applying the *h*-index function to each author ...

```{r apply-h-index}
h_indexes <-
  author_citations %>%
  group_by(author) %>%
  summarise(h_index_inc_self = h_index(citations_inc_self),
            h_index_exc_self = h_index(citations_exc_self)) %>%
  mutate(rank_inc_self = min_rank(-h_index_inc_self),
         rank_exc_self = min_rank(-h_index_exc_self)) %>%
  arrange(rank_inc_self, rank_exc_self, author)
```

Job done.  Before revealing the top authors, here are the ones that lose the
most by discounting self-citations.  Richard Cotton is the only person to lose
more than one *h*-index point.

```{r self-citation-losers}
# Biggest losses to excluding self-citations
h_indexes %>%
  dplyr::filter(h_index_exc_self != h_index_inc_self) %>%
  mutate(h_index_loss = h_index_inc_self - h_index_exc_self) %>%
  top_n(1, h_index_loss) %>%
  arrange(desc(h_index_loss), rank_inc_self, rank_exc_self, author) %>%
  knitr::kable()
```

Your reward for reading so far even though you'd already seen this graph
in the tweet.

```{r graph}
annotate_authors <- function(x, y, label) {
  annotate("text", x, y, angle = 90, hjust = 0, label = label)
}

h_indexes %>%
  rename(h_index = h_index_inc_self) %>%
  count(h_index) %>%
  arrange(h_index) %>%
  mutate(h_index = fct_inorder(as.character(h_index))) %>%
  ggplot(aes(h_index, n)) +
  geom_point() +
  geom_segment(aes(xend = h_index, yend = 1)) +
  geom_vline(xintercept = 15.5, linetype = 2, colour = "grey50") +
  scale_y_log10(limits = c(1, NA)) +
  annotation_logticks(sides = "l") +
  annotate_authors("39", 1.5, "Wickham") +
  annotate_authors("31", 1.5, "RStudio") +
  annotate_authors("19", 1.5, "Hornik") +
  annotate_authors("18", 3, "R Core, Maechler") +
  annotate_authors("14", 1.5, "Zeileis") +
  annotate_authors("13", 1.5, "Ripley") +
  annotate_authors("12", 6, "Allaire, Bolker, Chang, Xie") +
  xlab(expression(paste("CRAN ", italic("h"), "-index"))) +
  ylab("Number of authors") +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.ticks.x = element_blank()) +
  ggtitle(expression(paste("CRAN ", italic("h"), "-index by number of authors")))
```

A couple of points:

* The only women in at least the top 11 are employees of RStudio.
* In the CRAN context the *h*-index rewards packages that are used by other
  packages, so authors of foundational or 'back-end' packages are more likely to
  have high *h*-indexes than authors of presentational or 'front-end' packages.

Here are some authors with high numbers of downloads *of packages they maintain*
but *h*-indexes of zero.

<aside>

Downloads from the RStudio CRAN mirror on 13 February 2019.

</aside>

* [archdata](https://cran.r-project.org/package=archdata) is the only CRAN
  package maintained by [David L
  Carlson](https://cran.r-project.org/web/checks/check_results_dcarlson_at_tamu.edu.html).
  It was downloaded 1700 times on 13 February 2019, but isn't depended on or
  imported by any other CRAN packages, perhaps because it provides data rather
  than functions.  Carlson has no other CRAN packages.
* [prophet](https://cran.r-project.org/web/packages/prophet/index.html) is the
  only CRAN package maintained by [Sean
  Taylor](https://cran.r-project.org/web/checks/check_results_sjt_at_fb.com.html).
  It isn't depended on or imported by any other CRAN packages even though it
  provides functions, perhaps because its function isn't primarily for
  developers.

```{r download-downloads, include = FALSE, eval = FALSE}
download.file("http://cran-logs.rstudio.com/2019/2019-02-13.csv.gz",
              here("downloads.csv.gz"))
```

```{r parse-downloads, include = FALSE}
downloads <- read_csv(here("downloads.csv.gz"))
```

```{r join-downloads, echo = FALSE}
package_maintainers <-
  pkglist %>%
  select(package, maintainer) %>%
  mutate(maintainer = clean_up_author(maintainer)) %>%
  unnest()

downloads %>%
  count(package) %>%
  rename(downloads = n) %>%
  inner_join(package_maintainers, by = "package") %>%
  group_by(maintainer) %>%
  summarise(downloads = sum(downloads)) %>%
  inner_join(h_indexes, by = c("maintainer" = "author")) %>%
  dplyr::filter(h_index_inc_self == 0) %>%
  arrange(desc(downloads)) %>%
  select(maintainer, downloads)
```

## Your CRAN *h*-index

Search the full table for your own *h*-index.

```{r datatable, echo = FALSE}
datatable(h_indexes)
```

<details><summary>Expand for session info</summary>

```{r session-info, tidy = FALSE}
options(width = 80)
session_info()
```

</details>
