---
title: "R rounding is weird?  Try JavaScript!"
description: |
  Reproducing browser rounding in R
author: Duncan Garmonsway
date: April 25, 2016
preview: "rounding.png"
output:
  radix::radix_article:
    self_contained: false
resources:
  exclude:
    data
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "##", collapse = TRUE, cache = TRUE)
```

## round(0.5) == 0?  Eh?

A common source of confusion in R is rounding-to-even (example adapted from
?round):

```{r rounding-confusion, echo = TRUE)
matrix(c(x1, round(x1)), nrow = 2, byrow = TRUE) #-- IEEE rounding !
```

This post does five things:

* Illustrates rounding bias with a graph.
* Discovers that JavaScript (in my browser) rounds the other way.
* Encounters floating-point difficulties when emulating JavaScript.
* Calls JavaScript itself via the `V8` package.
* Explains where all the time goes.

## Graph of rounding bias

Here is an unpolished graphical illustration of the bias introduced by rounding
halves (0.5, 1.5, etc.) away from zero.  The details of the difference are
neatly explained in the [R
Inferno](http://www.burns-stat.com/pages/Tutor/R_inferno.pdf), circle 8.1.52.

```{r rounding-to-even}
# Round-to-even is the R default
round_to_even <- round

```{r rounding-away-from-zero}
# Round-away-from-zero function adapted from http://stackoverflow.com/a/12688836/937932
round_away_from_zero = function(x, digits = 0) {
  posneg = sign(x)
  z = abs(x) * 10^digits
  z = z + 0.5
  z = trunc(z)
  z = z * 10^(-digits)
  z*posneg
}
```

```{r rounding-bias-1, echo = FALSE}
# Compare the two systems
# x1 <- seq(-2.5, 2.5)
# matrix(c(x1,
#          round_to_even(x1),
#          round_to_odd(x1)),
#        nrow = 3, byrow = TRUE)

# Graph the bias of many random draws
N <- 10000
bias<- function(FUN) {
  # Round to one decimal place to ensure 0.5 ever appears.
  # Draw between 0 and 2 to fairly represent both 0.5 and 1.5.
  x <- round(runif(10, min = 0, max = 2), 1)
  mean(FUN(x) - x)
}

bias_to_even <- replicate(N, bias(round_to_even))
bias_away_from_zero <- replicate(N, bias(round_away_from_zero))

limits <- c(-0.5, 0.5)
par(mfrow = c(2, 1))
  hist(bias_to_even, xlim = limits, col = "lightgreen")
  hist(bias_away_from_zero, xlim = limits, col = "lightblue")
```

## JavaScript rounding in the Chrome browser

But when I tried to emulated a website's behaviour in R, it turned out that
Chrome was rounding towards odd numbers after the decimal point (anyone know
why?).  Try the following in the Chrome Developer Console (ctrl+shift+c in a
tab).

```js
// Rounds away from zero
console.log((-1.5).toFixed(0));
console.log((-0.5).toFixed(0));
console.log((0.5).toFixed(0));
console.log((1.5).toFixed(0));

// Rounds to odd
console.log((-0.25).toFixed(1));
console.log((-0.15).toFixed(1));
console.log((-0.05).toFixed(1));
console.log((0.05).toFixed(1));
console.log((0.15).toFixed(1));
console.log((0.25).toFixed(1));
```

How odd is that?  So I adapted a [handy MATLAB
implementation](https://www.mathworks.com/matlabcentral/fileexchange/40286-rounding-functions-collection)
of rounding-to-odd, and compared it with the other two strategies.

```{r rounding-to-odd}
# Round-to-odd function adapted from
# https://www.mathworks.com/matlabcentral/fileexchange/40286-rounding-functions-collection
round_to_odd <- function(x, digits = 0) {
  y <- (x) * 10^digits
  z <- y %% (2 * sign(y))
  z[is.nan(z)] <- 0
  z[abs(z) != 1.5] <- 0
  z <- round_away_from_zero(y - z/3)
  z * 10^(-digits)
}
```

```{r rounding-bias-2, echo = FALSE}
# Compare the three systems
# x1 <- seq(-2.5, 2.5)
# matrix(c(x1,
#          round_to_even(x1),
#          round_away_from_zero(x1),
#          round_to_odd(x1)), nrow = 4, byrow = TRUE)

bias_to_odd <- replicate(N, bias(round_to_odd))

limits <- c(-0.5, 0.5)
par(mfrow = c(3, 1))
  hist(bias_to_even, xlim = limits, col = "lightgreen")
  hist(bias_away_from_zero, xlim = limits, col = "lightblue")
  hist(bias_to_odd, xlim = limits, col = "pink")
```

The graph shows that, like rounding-to-even, rounding-to-odd is unbiased, but a
snag is that successive rounded operations will never reach zero (see comments
on [this StackExchange
answer](http://programmers.stackexchange.com/a/256269/111311)):

```{r rounding-reach-zero, echo = TRUE}
x <- 77
for(i in 1:10) {
  x <<- round_to_odd(x/2)
  cat(x, ", ")
}

x <- 77
for(i in 1:10) {
  x <<- round_to_even(x/2)
  cat(x, ", ")
}
```

## Floating point errors

Using my new round-to-odd function to emulate JavaScript behaviour, I
encountered floating point errors.  For example, take the number 6.65:

```{r rounding-floatingpoint1, echo = TRUE}
sprintf("%.16f", c(6.65, 7 * 0.95, 7 - 0.35))
```

The tiny differences don't affect rounding in R:
```{r rounding-floatingpoint2, echo = TRUE}
round_to_odd(c(6.65, 7 * 0.95, 7 - 0.35), 1)
```

But they do affect rounding in JavaScript.  Again, paste these into the browser
console:

```
console.log((6.65).toFixed(1));
console.log((7 * 0.95).toFixed(1));
console.log((7 - 0.35).toFixed(1));
```

## Calling JavaScript V8 engine via the V8 package

At this point, I gave up on emulating JavaScript behaviour in R, and resorted to
calling JavaScript from R via the
[`V8`](https://cran.r-project.org/web/packages/V8/index.html) package, which uses
the V8 JavaScript engine, the same that my browser (Chrome) uses.

```{r rounding-loadV8, echo = TRUE}
library(V8)
```

```{r rounding-V8, echo = TRUE}
# library(V8)
ct <- V8::v8()
roundjs <- function(x, digits) {
  sapply(x, function(y) {ct$get(paste0("Number((", sprintf("%.16f", y), ").toFixed(", digits, "))"))})
}
roundjs(c(6.65, 7 * 0.95, 7 - 0.35), 1)
```

## What took me so long

This was a particularly tricky part of a bigger project (see next week's
post).


Most of the time went on finding, testing and correcting the two rounding
functions for round-to-odd and round-away-from-zero.  I adapted the round-to-odd
function from some [handy MATLAB
implementations](https://www.mathworks.com/matlabcentral/fileexchange/40286-rounding-functions-collection)
of various rounding strategies.  Unfortunately, they depended on MATLAB's
built-in `round` function, which, according to its
[documentation](https://uk.mathworks.com/help/matlab/ref/round.html), rounds
away from zero, so I had to find a round-away-from-zero function in R first.
Even then, it didn't work for negatives when I ported it to R, probably due to
fundamental language differences:

```{r rounding-surprising, echo = TRUE}
# Surprising behaviour of
-1.5 %% 2
# Predictable behaviour (but different to MATLAB?)
-1.0 %% 0
```

I also spent quite a while on the graphs of bias, where I befuddled myself by
drawing random numbers between 0 to 1 (which is unfair on unbiased functions,
because only 0.5 is represented, not 1.5), and by not doing preliminary rounding
on the random draws (which meant that 0.5, 1.5, etc., weren't represented at
all).

Finally, my initial `V8` function used the `V8` package's own magic for passing
values to the V8 engine, but when it didn't work, I suspected that the values
were being passed as a string, and that R was rounding them as part of the
conversion.  For example:

```{r rounding-V8old, echo = TRUE}
library(V8)
roundjs <- function(x, digits) {
  ct <- V8::v8()
  ct$source(system.file("js/underscore.js", package="V8")) # Essential for _
  ct$assign("digits", digits)
  xrounded <-
    ct$call("_.forEach",
            x,
            V8::JS("function(item, index, arr) {arr[index] = Number(item.toFixed(digits));}"))
  xrounded
}
roundjs(c(6.65, 7 * 0.95, 7 - 0.35), 1)
```

## Code for the graphs

```{r rounding-recall1, ref.label = 'rounding-bias-1', echo = TRUE, eval = FALSE}
```

```{r rounding-recall2, ref.label = 'rounding-bias-2', echo = TRUE, eval = FALSE}
```

