---
title: "How many StatsBloggers are there?"
author: Duncan Garmonsway
description: |
  Parsing the daily digest emails and thinking about modelling
date: May 2, 2016
preview: "statsblogs.png"
output:
  radix::radix_article:
    self_contained: false
resources:
  exclude:
    data
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "##", collapse = TRUE, cache = TRUE)
```

This post does two things:

* Reproduces my [R-Bloggers
  post](https://nacnudus.github.io/duncangarmonsway/posts/2016-04-18-rbloggers/)
  on the [StatsBlogs](http://www.statsblogs.com) website, to discover how many
  StatsBlogs blogs there are *really*.
* Considers how to interpret changes in diversity without knowingly observing
  births/deaths of blogs.

Many people who blog about statistics syndicate their posts on the the
[StatsBlogs](http://www.statsblogs.com) website by the [Talk Stats
forum](http://www.talkstats.com).  The list of "contributing blogs" tends only
to lengthen, but how many actual posts are there in a given week/month, from how
many different blogs?

## The gist of it

I subscribed to the [StatsBlogs](http://www.statsblogs.com/) daily digest emails
in February 2014, giving me a good time-series of posts.  See my [R-Bloggers
post](`r htmltools::HTML('{{< relref "2016-04-18-rbloggers.html" >}}')`) and the
code at the end of this post for how I mined the emails for names and dates.

## The trends

It turns out that there are have been about 30 blogs active in a given month,
posting about 150 posts (the only one that regularly posts more than once per
week is, no prizes for guessing, [Statistical Modeling, Causal Inference, and
Social Science](http://andrewgelman.com/)).  There was a change in mid-2015,
either a step-change down from ~175 blogs/month, or the start of a decline.
It's hard to say which.  When I first subscribed in February 2014, there were
over 200 posts per month.  Please comment if you can suggest reasons for the
change.

```{r statsblogs-packages, echo = FALSE}
library(dplyr)
library(tidyr)
library(tm.plugin.mail)
library(xml2)
library(stringr)
library(lubridate)
library(ggplot2)
library(scales)
library(fs)
library(here)
```

```{r statsblogs-wrangle, echo = FALSE}
# TODO: Implement the following manual steps in R
# 1. Create a label in Gmail for the R-Bloggers digests
# 2. Export emails with that label from Gmail
# 3. Extract the exported file
# 4. Run sed ':a;N;$!ba;s/=\r\n//g' statsblogs in the Mail directory, i.e.
#    sed ':a;N;$!ba;s/=\r\n//g' ./Takeout/Mail/statsblogs > ./clean.mbox
#    This steps removes funny newlines that are in the original emails in my
#    inbox.

mail_dir <- here("data", "temp")

# Extract the emails into individual files
convert_mbox_eml(here("data", "clean.mbox"), mail_dir)

# Collect the file names
emails <- dir_ls(mail_dir, full.names = TRUE)

# Collect the datetimes in the first line of each file
# Also collect the journals from the subject lines
n <- length(emails)
datetimes <- vector("character", length = n)
blogs  <- vector("list", length = n)
i <- 0
for (filename in emails) {
  i <<- i + 1
  datetimes[i] <- readLines(filename, n = 1)
  # Extract the links to the original blogs
  blogs[[i]] <-
    read_html(filename) %>%
    xml_find_all("//div/p[(@class = '3D\"syndicated-attribution\"')][1]//a[1]") %>%
    xml_text
}

# Extract the datetime string
datetimes <-
  datetimes %>%
  str_sub(start = 34) %>%
  strptime(format = "%b %d %H:%M:%S %z %Y")

# Link the datetime with individual blogs
names(blogs) <- datetimes
blogs <- stack(blogs)

# Recover the dates and clean the blog names
blogs <-
  blogs %>%
  rename(blog = values, datetime = ind) %>%
  mutate(datetime = ymd_hms(datetime),
         # blog = str_replace(blog, fixed("=\\n"), ""),
         blog = str_replace(blog, fixed("=C2=BB"), "»"),
         blog = str_replace(blog, fixed("=E2=80=93"), "–"),
         blog = str_replace(blog, fixed("=E2=80=A6"), "…"),
         blog = str_replace(blog, fixed("=C3=A6"), "æ"),
         blog = str_replace(blog, fixed("=EA=B0=84=EB=93=9C=EB=A3=A8=EB=93=9C =ED=81=AC=EB=A6=AC=EC=8A=A4=ED=86=A0=ED=8C=8C"), "(간드루드 크리스토파)"),
         blog = str_replace(blog, fixed("=D0=AF=D1=82=D0=BE=D0=BC=D0=B8=D0=B7=D0=BE"), "Ятомизо"),
         blog = str_trim(blog))

# Number of different blogs per week/month/year
periodically <- function(x, period) {
  x %>%
    group_by_(period, "blog") %>%
    summarise(posts = n()) %>%
    group_by_(period) %>%
    summarise(posts = sum(posts),
              blogs = n()) %>%
    gather(measure, n, posts, blogs) %>%
    rename_("date" = period) %>%
    mutate(period = paste0(period, "ly"))
}
periodical <-
  blogs %>%
  mutate(year = ceiling_date(datetime, "year"),
         month = ceiling_date(datetime, "month"),
         week = ceiling_date(datetime, "week"))
monthly <- periodically(periodical, "month")
weekly <- periodically(periodical, "week")
periodical <-
  bind_rows(monthly, weekly) %>%
  group_by(period) %>%
  filter(date > min(date),
         date < max(date)) %>%
  ungroup
```

```{r statsblogs-periodical, echo = FALSE}
periodical %>%
  ggplot(aes(date, n, colour = period, linetype = measure)) +
  geom_line() +
  xlab("") +
  ylab("Number of blogs/posts") +
  ggtitle("Number of StatsBlogs blogs/posts per week/month")
```

```{r statsblogs-top10, echo = FALSE}
# Most prolific bloggers
blogs %>%
  count(blog) %>%
  arrange(desc(n)) %>%
  slice(1:10) %>%
  mutate(blog = str_wrap(blog, 30)) %>%
  mutate(blog = factor(blog, levels = blog)) %>%
  ggplot(aes(blog, n, label = blog)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  xlab("") +
  ylab("Number of posts since February 2014") +
  ggtitle("Top-10 bloggers by number of posts since February 2014") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r statsblogs-rates, echo = FALSE}
# Rates of blogging (smoothed difference in days)
blogs %>%
  count(blog, datetime) %>%
  group_by(blog) %>%
  arrange(datetime) %>%
  mutate(days = difftime(datetime, lag(datetime), units = "days"),
         days = as.numeric(days),
         days = days + .1) %>%         # Avoid log(0) errors
  filter(n() >= 85) %>%
  ungroup %>%
  ggplot(aes(datetime, days)) +
  geom_point(colour = "grey") +
  geom_smooth(se = FALSE) +
  scale_y_continuous(breaks = c(7, 28)) +
  coord_cartesian(ylim = c(0, 35)) +
  scale_x_datetime(breaks = date_breaks("6 months"), labels = date_format("%b'%y")) +
  facet_wrap(~blog, label = label_wrap_gen()) +
  xlab("") +
  ylab("Days between posts") +
  ggtitle("Days between posts on most-prolific blogs") +
  theme(legend.position = "none")
```

## Survival modelling without birth/death observations

Can we do a survival analysis without knowingly observing births and deaths of
blogs?  I haven't trawled the blogs to find their first-ever posts, and it would
be hard even for an author to identify a last-ever post.  Without that crucial
information, I doubt a hazard function can be estimated, though I don't know an
awful lot about that kind of thing, so maybe.

But what about diversity?  I think we could get somewhere even without births
and deaths.  Here's the cumulative distribution of observed blogs (the number of
different blogs observed), over the whole period.

```{r statsblogs-cumdist}
blogs %>%
  # Get the first observation of each blog
  group_by(blog) %>%
  arrange(datetime) %>%
  slice(1) %>%
  ungroup %>%
  arrange(datetime) %>%
  mutate(cumulative = 1:n()) %>%
  ggplot(aes(datetime, cumulative)) +
  geom_line() +
  xlab("") +
  ylab("Number of different blogs observed") +
  ggtitle("Cumulative number of different blogs observed")
```

Supposing the population of blogs were static, then the first six months
of the cumulative distribution curve would make sense.  Lots of blogs post daily,
weekly or monthly, so by the time a couple of months have gone by, many blogs
have already been observed.  After that, things slacken, until after about six
months the curve levels off -- all blogs have been observed.

Except that it doesn't level off.  It continues to rise steadily, implying that
new blogs are being syndicated.  On the other hand, the number of different
blogs observed in a given month (first graph) is slowly declining, so some
blogs must be posting less often, or ceasing altogether.  Given those slopes,
the composition of the population must be changing.

I'm no clever clogs, so I'm not about to develop a statistic to describe those
two slopes, to figure out their distribution, or to test hypotheses.  If anyone
knows anything about this, please comment!

## What took so long

Almost nothing, since this post re-used the code from my [R-Bloggers post](`r
htmltools::HTML('{{< relref "2016-04-18-rbloggers.html" >}}')`).  All I had to
do was tweak the XPath, and then take a long bath to think about cumulative
distributions.

## Code

Nothing postworthy, so see
[GitHub](https://github.com/nacnudus/duncangarmonsway/blob/master/_posts/2016-05-02-statsblogs/2016-05-02-statsblogs.Rmd)
if you're interested.

