---
title: "Get me to the church on time with R spatial"
description: |
  Using Google location data to prove church attendance and get married.
author: Duncan Garmonsway
date: May 1, 2019
output:
  radix::radix_article:
    self_contained: false
resources:
  exclude: ["locationdata.Rds", "Location\\ History.json", "google-api-key"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, echo = TRUE, cache = TRUE)
```

![Tweet: "When my church asked for evidence of regular attendance, I bet they
didn't expect #rstats {sf} {ggcal} and Google location history." Reply by
@mattdray: "Ah, so you turned to script in liu of scripture."](./tweet.png)

Churches have excellent transport connections, plenty of seating, a resident
celebrant, are open on weekends, and are often pretty.  In other words they are
excellent wedding venues.  But they can, and probably will, decline to marry
couples that have no connection with the church, such as living in the parish,
and/or worshipping there.

My fiancée and I applied to marry at a church where neither of us lives, but
where I sing in the choir.  The church responded with a form for us to circle my
qualifying connection (regular worshipper for six months) and -- the
catch -- provide documentary evidence.

Evidence?

So I downloaded my Google location data.  It's always on (God and Google know
the inmost secrets of my heart), and like everyone else I keep my phone within
arm's reach 24/7.  All I needed then was:

* {jsonlite} to read the Google location data dump
* {ggmap} to check that it made sense
* {sf} to filter for points in a radius
* {ggcal} to mark the days on a calendar

There are a couple of similar blog posts around, so I won't dwell on the parts
that worked.

```{r library}
library(jsonlite)
library(lubridate)
library(tidyverse)
library(sf)
library(ggmap)
library(ggcal) # devtools::install_github("jayjacobs/ggcal")
library(here)
```

Where is the church?  Google knows.

```{r church}
church <-
  st_sf(name = "church", st_sfc(st_point(c(55.8735208,-4.2763132)))) %>%
  st_set_crs(4326) # WGS84 (GPS)
```

Where was I?  Google knows.  For this post I load it from an R data file because
reading the JSON is slow.

```{r load_data}
# locationdata <- fromJSON(here("Location History.json"))
# saveRDS(locationdata, "locationdata.Rds")
locationdata <- readRDS(here("locationdata.Rds"))
```

This is the jankey bit.  Google's latitude and longitude are integers, so you
have to divide them by ten million (fine) but the longitude is wrong by 69.4970
degrees or so.  Perhaps it isn't wrong for you.  I also had to take the modulus
because some points have wrapped around the earth to more than 360.  I, um,
Googled the problem but couldn't _locate_ an explanation.  The final line
calculates the distance of each point from the church, in meters or feet or
seconds or something.

```{r locations}
locs <-
  locationdata$locations %>%
  as_tibble() %>%
  select(latitudeE7, longitudeE7, `timestampMs`) %>%
  transmute(lat = latitudeE7 / 1E7,
            lon = (longitudeE7 / 1E7) %% 360 - 69.4970, # Google you had ONE JOB
            date = as.POSIXct(as.numeric(timestampMs)/1000, origin="1970-01-01"),
            day = as.Date(date)) %>%
  st_as_sf(coords = c("lat", "lon"), remove = FALSE) %>%
  st_set_crs(4326) %>% # WGS84 (GPS)
  mutate(dist_from_church = as.numeric(st_distance(geometry, church)))
```

To show you where the magic number 69.4970 came from, here's a map of the
points.  I refined the magic number until they lined up with the roads.

```{r map}
register_google(key = read_file(here("google-api-key")))

m <- get_map(location = c(lat = 55.8735208, lon = -4.2763132), # church again
             source = "stamen",
             zoom = 15)

ggmap(m) +
  geom_point(aes(-4.2763132, 55.8735208)) +
  geom_point(aes(lon, lat),
             colour = "red",
             data = filter(locs, dist_from_church < 1000))
```

{ggcal} plots calendars as graphs.  This is what went to the vicar.  Not really,
because I'm not getting married in Glasgow, hah tricked you, but I did used to
sing there a while back so it makes a safe-ish example for this post, and I did
send the real one to the parish office.

```{r}
in_church <-
  locs %>%
  filter(dist_from_church <= 100) %>% # So what if I am in the tea shop?
  filter(date <= ymd("2017-02-28")) %>%
  distinct(day) %>%
  pull(day)

in_church_range <- range(in_church)
in_church_cal <- seq(in_church_range[1], in_church_range[2], by = "day")
in_church_today <- in_church_cal %in% in_church

ggcal(in_church_cal, in_church_today) +
  ggtitle("Days when Duncan Garmonsway was within 100 meters of church",
          subtitle = "Google location history")
```

<details><summary>Expand for session info</summary>

```{r session_info}
options(width = 100)
sessioninfo::session_info()
```

</details>

