---
title: "Import ragged data with readr::melt_csv()"
description: |
  Melt it down and start again
author: Duncan Garmonsway
date: January 14, 2019
preview: "im_melting_wicked_witch_of_the_west.jpg"
output:
  radix::radix_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, include = FALSE}
library(tidyverse)
library(here)
```

Standard tools like `readr::read_csv()` can cope to some extent with unusual
inputs, like files with empty rows or newlines embedded in strings.  But some
files are so whacky that standard tools don't work at all, and instead you have
to take the file to pieces and reassemble it in a standard design.

The [*readr*](https://readr.tidyverse.org) package has [recently
acquired](https://www.tidyverse.org/articles/2018/12/readr-1-3-1/) a set of
tools for taking a file to bits.  They are the
[`melt_*()`](https://readr.tidyverse.org/reference/melt_delim.html) family.

* [`melt_csv()`](https://readr.tidyverse.org/reference/melt_delim.html), [`melt_csv_chunked()`](https://readr.tidyverse.org/reference/melt_delim.html)
* [`melt_csv2()`](https://readr.tidyverse.org/reference/melt_delim.html), [`melt_csv2_chunked()`](https://readr.tidyverse.org/reference/melt_delim.html)
* [`melt_tsv()`](https://readr.tidyverse.org/reference/melt_delim.html), [`melt_tsv_chunked()`](https://readr.tidyverse.org/reference/melt_delim.html)
* [`melt_delim()`](https://readr.tidyverse.org/reference/melt_delim.html), [`melt_delim_chunked()`](https://readr.tidyverse.org/reference/melt_delim.html)
* [`melt_table()`](https://readr.tidyverse.org/reference/melt_table.html), [`melt_table2()`](https://readr.tidyverse.org/reference/melt_table.html)
* [`melt_fwf()`](https://readr.tidyverse.org/reference/melt_fwf.html)

<aside>
![Take it to pieces and start again.  Rubik's Cube disassembled, from
[Wikipedia](https://en.wikipedia.org/wiki/File:Disassembled-rubix-1.jpg).](rubik-cube-disassembled.jpg)
</aside>

The `melt_*()` family separates delimited text files into individual cells.  So
"melt" isn't quite the right name -- it should be "disassemble" because it's
about separating the pieces, but "melt" is much shorter to type

<aside>
The term "melt" also has heritage from [`reshape2::melt()`](https://github.com/hadley/reshape2), although it doesn't do quite the same thing.
</aside>

## The problem with non-rectangular data

Here's a contrived example that breaks two assumptions made by common tools like
`readr::read_csv()`.

1. There are more cells in some rows than others.
1. There are mixed data types within each column.

```
Help,,007,I'm
1960-09-30,FALSE,trapped in,7,1.21
non-rectangular,data,NA
```

```{r help-im-trapped-data, include = FALSE}
data <-
"Help,,007,I'm
1960-09-30,FALSE,trapped in,7,1.21
non-rectangular,data,NA"
```

To force that data into a data frame, `readr::read_csv()` must make drastic
adjustments.

1. It omits the cell "1.21".
1. It invents an `NA` cell to fill a gap (row 3 column 4).
1. It promotes the data type of every cell to character so that it can combine
   them in columns.

```{r read-csv}
readr::read_csv(data, col_names = FALSE)
```

Omission of cells is a big problem because it could occur silently.  You can
force `readr()` to read the `1.21` cell by specifiying column names for five
columns, but then it has to invent `NA` cells to plug the gaps in the fifth
column.

```{r read-csv-colnames}
readr::read_csv(data, col_names = letters[1:5])
```

The invention of `NA` cells is a problem when you need to know the difference
between a cell being missing altogether, and a cell that contains a 'missing'
value.  This is a bit niche, but a short row (fewer cells) could mark a
different table within the same file (it happens), whereas `NA` means that a
value might have existed yet wasn't recorded.

Type promotion is the least of the problems because it's lossless -- the file is
a text file in the first place.  But it would be nice to traverse the cells and
pluck out, say, any dates.

## Introducing `readr::melt_csv()`

Now see what `melt_csv()` does.  Remember the csv file looks like this.

```
Help,,007,I'm
1960-09-30,FALSE,trapped in,7,1.21
non-rectangular,data,NA
```

The `melt_csv()` function reads the file one cell at a time, importing each cell
of the file into a whole row of the final data frame.

```{r melt-csv}
readr::melt_csv(data)
```

Think of this output like painting-by-numbers.  In the same way that a
painting-by-numbers stencil *describes* a painting (the colour, shape and
position of each brush-stroke), this 'melted' form *describes* a CSV file (the
value, datatype and position of each cell).  While a painting isn't structured
data, a painting-by-numbers stencil *is* structured data about a painting, and
while a CSV file isn't necessarily structured data, a melted CSV *is* structured
data about a CSV file.

<aside>
You might have seen 'melted' data before in [David Robinson's
article](http://rpubs.com/dgrtwo/tidying-enron), the
[tidyr](https://nacnudus.github.io/tidyr) and
[unpivotr](https://nacnudus.github.io/unpivotr) documentation, or [Spreadsheet
Munging Strategies](https://nacnudus.github.io/spreadsheet-munging-strategieps).
</aside>

![Paint-by-numbers image of a house in the countryside, before and after
painting.  Image from from *Paint by Number: The How-To Craze that Swept the
Nation* by William L. Bird, via
https://simanaitissays.com/2014/09/05/by-the-numbers/.](./paint-by-numbers.jpg)

What the `melt_csv()` output means:

* The output of `melt_csv()` is a data frame -- *structured* data about
  *un*-structured data!
* Each row of the data frame represents one cell of the input data.
* The empty cell of row 1 is imported, but the *missing* cells at the ends of
  rows 1 and 3 aren't.
* No data type conversion is attempted -- every value is imported as a string,
  and the `data_type` column merely gives *readr*'s best guess of what the data
  types ought to be.

How can you use this?  To begin with, you can do some simple manipulations with
ordinary functions.  For example you could extract the words.

```{r filter-for-character}
data <- "Help,,007,I'm
1960-09-30,FALSE,trapped in,7,1.21
non-rectangular,data,NA"

readr::melt_csv(data) %>%
  dplyr::filter(data_type == "character")
```

Was any of the message lost in transmission?  Yes, two segments are missing.

```{r filter-for-missing}
readr::melt_csv(data) %>%
  dplyr::filter(data_type == "missing")
```

## Using melted data

Use-cases are hard to come by because this is a (relatively) new way to work
with messy data.  Try out [New Zealand's election
statistics](https://www.electionresults.org.nz/electionresults_2017/statistics/csv/candidate-votes-by-voting-place-1.csv)
or some [UK civil servants'
expenses](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/413595/April-june-2014.csv)
for yourself, or [watch me wrangle](https://www.youtube.com/watch?v=zUqwJuZPPPs)
a [#TidyTuesday](https://github.com/rfordatascience/tidytuesday) dataset of US
honey production with the help of my
[unpivotr](https://nacnudus.github.io/unpivotr) package.

My [tidyxl](https://nacnudus.github.io/tidyxl) package melts spreadsheets in a
similar way, and [Spreadsheet Munging
Strategies](https://nacnudus.github.io/spreadsheet-munging-strategies/) has many
worked examples and case studies, most of which could apply to plain text files.

![The Wicked Witch of the West, melting, from the film The Wizard of
Oz.](im_melting_wicked_witch_of_the_west.jpg)
