---
title: "With added bacran"
description: |
  Nerd-sniped by kevinbacran, a graph of CRAN co-authorship.
author: Duncan Garmonsway
date: March 03, 2019
output:
  distill::distill_article:
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)

library(kevinbacran)
library(tidyverse)
library(tidygraph)
library(igraph)
library(ggraph)
```

[Matt Dray](https://twitter.com/mattdray) came up with the [hilarious
pun](https://www.rostrum.blog/2019/02/27/hadley-number/) 'kevinbacran' and then
had to follow through and make the
[package](https://github.com/matt-dray/kevinbacran).  It calculates the degrees
of separation between CRAN package authors and Hadley Wick*ham*.  Return here
when you've [played with the shiny
app](https://mattdray.shinyapps.io/hadley-number/).

<aside>
Pun included for compliance with the kevinbacran terms and conditions of reuse.
</aside>

If you're still reading this then you have been nerd-sniped, as I was.  To help
you end your procrastination quickly, here are answers to a few questions:

1. Who has the highest Hadley number?
1. What is the longest 'shortest path' between any two CRAN authors?
1. What is the largest network disconnected from Hadley?
1. Is Hadley the most central author?

## 0. Preparation

The `kevinbacran` package provides a graph of package co-authorship, so I piggy
back on that, just as it is my pleasure to do with Matt's professional work.
Nodes are authors; edges are packages.

```{r load-graph}
cran_pkg_graph
```

Incredibly, the  [igraph](https://cran.r-project.org/package=igraph) package is
willing and able to calculate the length of the shortest route between all
nodes, all at once, in a few seconds.  It returns a square matrix, a bit like
old people used to have on the back of a road atlas, with the driving times
between cities, except here it's the number of people in the chain linking one
author with another.  Infinite values represent impossible connections, but I
subsituted `NA` for these so that they don't upset subsequent `max()` and
`min()`.

```{r max-distances, cache = FALSE}
cran_pkg_igraph <- as.igraph(cran_pkg_graph)
dists <- distances(cran_pkg_igraph)
dists[is.infinite(dists)] <- NA # is there a more efficient way?
```

## 1. Who has the highest Hadley number?

First I need to know which column (or row) represents Hadley.  Then I can search
that row (or column) for the highest value.

```{r highest-hadley}
# The row and column number of Hadley Wickham in the matrix of distances.
hadley <-
  cran_pkg_graph %>%
  as_tibble() %>%
  dplyr::filter(name == "Hadley Wickham") %>%
  pull(number)
hadley

# The row of distances between Hadley and everyone else.
hadley_row <- dists[hadley, ]

# The maximum distance between Hadley and anyone else.
(max_hadley_row <- max(hadley_row, na.rm = TRUE))

# The people at the greatest distance from Hadley.
which(hadley_row == max_hadley_row)
```

Congratulations Chelsey, Nicholas and Franz, you are Hadley's most distant
relations.

## 2. What is the longest 'shortest path' between any two CRAN authors?

I now wipe out the entire upper triangle (and the diagonal), because I don't
want to notice each path twice.  This question is undirected.  I only need the
maximum value in the bottom triangle.

```{r all-most-distant-authors, cache = FALSE}
dists[upper.tri(dists, diag = TRUE)] <- NA # Can it be done faster?

# Maximum distance beteween any two authors
(maximum_distance <- max(dists, na.rm = TRUE))

max_distances <-
  which(dists == maximum_distance, arr.ind = TRUE) %>%
  as_tibble() %>%
  mutate(from = rownames(dists)[row],
         to = colnames(dists)[col],
         distance = map2_int(row, col, ~ as.integer(dists[.x, .y]))) %>%
  select(from, to, distance)
max_distances
```

Oh hi Chelsey, Nicholas and Franz! Not only are you Hadley's most distant
relations, you are among the most distantly related package authors in the whole
graph.  What did you write?

```{r distant-authors-packages}
nodes <- as_tibble(activate(cran_pkg_graph, nodes))
edges <- as_tibble(activate(cran_pkg_graph, edges))

nodes %>%
  dplyr::filter(name %in% c("Chelsey Bryant",
                            "Nicholas R Wheeler",
                            "Franz Rubel")) %>%
  inner_join(edges, by = c("number" = "from")) %>%
  distinct(name, package)
```

All three collaborated on the same package:
[kgc](https://cran.r-project.org/package=kgc).  The three authors on the other
side of the relation are not collaborators -- they each authored a different
package.

```{r counterparts-packages}
nodes %>%
  dplyr::filter(name %in% c("Qihuang Zhang",
                            "Natalia Vilor",
                            "Di Shu")) %>%
  inner_join(edges, by = c("number" = "from")) %>%
  distinct(name, package)
```

A visualisation would be easier to interpret.  A few incantations are required:

1. Combine the six author names
1. Calculate the graph between each pair with `kevinbacran::kb_pair()`.
1. Merge and distinctify the graphs with `igraph::simplify()`.
1. Plot with `kevinbacran::kb_plot()`.

(As you can tell, I do very little of the work around here.)

```{r visualisation, fig.width = 9.5, fig.height = 7}
set.seed(2019-03-03)

max_distances %>%
  mutate(kb_pair = map2(from, to, kb_pair, tidy_graph = cran_pkg_graph)) %>%
  pull(kb_pair) %>%
  reduce(graph_join, by = c("name", "number", ".tidygraph_node_index")) %>%
  as.igraph() %>%
  igraph::simplify(remove.multiple = TRUE,
                   remove.loops = FALSE,
                   edge.attr.comb = "random") %>%
  as_tbl_graph() %>%
  kb_plot()
```

It's lovely to see that Ernst Wit and Ernst C Wit are such near relations in
work as in life.

<aside> All this co-authorship analysis depends on the thorough and marvellous
work of the [cranly](https://cran.r-project.org/package=cranly) package author,
Ioannis Kosmidis.  It isn't perfect. It isn't possible to be perfect with data
that isn't used at the source.
</aside>

Are many authors so distantly related?  What is the distribution of maximum
distance between authors?  That is, for each author, the maximum distance
between them and anyone else, what is the distribution of that value across all
authors?

```{r max-distance-histogram}
rowmaxes <- apply(dists, 1, max, na.rm = TRUE)
hist(rowmaxes)
```

I ... did not expect a bimodal distribution.  That's how naive I am about
graphs.

## 3. What is the largest network disconnected from Hadley?

The graph can be split into cliques -- my name for graphs that aren't connected
with one another.  Most cliques only have one author in them -- authors who
haven't collaborated with anyone.  But the second largest clique only has 29
people in it, whereas the largest clique has 9251 in it.

```{r cliques}
cliques <- components(cran_pkg_igraph)
table(cliques$csize)
```

Like nearly everyone else, Hadley Wickham is in the largest clique.  Who is in
the next largest clique?

```{r not-hadley, cache = TRUE}
hadley_clique <- cliques$membership["Hadley Wickham"]
max_non_hadley_size <- max(cliques$csize[-hadley_clique])

# The largest graph not connected with Hadley
largest_non_hadley_cliques <- which(cliques$csize == max_non_hadley_size)

# Members of those cliques
non_hadley_members <-
  cliques$membership[cliques$membership %in% largest_non_hadley_cliques]

# Graph of the cliques
cliques_graph <-
  cran_pkg_graph %>%
  activate(nodes) %>%
  filter(name %in% names(non_hadley_members))

# Packages authored in the largest non-Hadley cliques
pkgs <- tools::CRAN_package_db()

cliques_graph %>%
  activate(edges) %>%
  as_tibble() %>%
  distinct(package) %>%
  inner_join(pkgs[, c("Package", "Description")],
             by = c("package" = "Package")) %>%
  mutate(Description = str_replace_all(Description, "[\n ]+", " ")) %>%
  print(n = Inf)

set.seed(2019-03-03)
kb_plot(cliques_graph)
```

That is a clique of regression packages.

## 4. Is Hadley the most central author?

Uh, what even is [centrality](https://schochastics.github.io/netrankr/)?
~~Picking `?centrality_information` at random:~~

> ~~‘centrality_information’: centrality based on inverse sum of resistance~~
> ~~distance between nodes (‘netrankr’)~~

Crikey that takes a while, let's try
[`centrality_betweenness`](https://igraph.org/r/doc/betweenness.html).

```{r hadley-centrality}
centrality <-
  cran_pkg_graph %>%
  activate(nodes) %>%
  mutate(importance = centrality_betweenness(directed = FALSE))

centrality %>%
  activate(nodes) %>%
  as_tibble() %>%
  arrange(desc(importance))
```

Yes.
