---
title: "Harry Potter and the N-Grams of Sentiment"
description: |
  *Harry Potter* cliches and story arcs by n-grams and sentiment analysis
author: "Duncan Garmonsway"
date: July 13, 2016
preview: "harry-potter-sentiment.png"
output:
  radix::radix_article:
    self_contained: false
resources:
  exclude:
    data
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "##", collapse = TRUE, cache = TRUE)
```

```{r harry-potter-sentiment-packages, echo = FALSE}
# Run the script "data/harry-potter-sentiment.R" first.
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(directlabels)
library(tokenizers)
library(stringi)
library(here)
```

This post applies [Julia Silge](http://juliasilge.com/blog/)'s amazing story-arc
[sentiment](http://juliasilge.com/blog/Life-Changing-Magic/)
[analyses](http://juliasilge.com/blog/Life-Changing-Magic/) to the *Harry
Potter* books.

It also busts the myth that "turned on his heel" is the series' most common
phrase.

Here is a related [shiny
app](https://duncan-garmonsway.shinyapps.io/harry-potter) to explore the ideas
futher.

## Super-easy text-munging

Casual text-munging is no longer a pain, thanks to a couple of new packages,
[tidytext](https://github.com/juliasilge/tidytext) and
[tokenizers](https://github.com/lmullen/tokenizers), and a not-so-new one,
[stringi](https://github.com/gagolews/stringi).

When I last analysed Harry Potter a few years ago, the
[tm](https://cran.r-project.org/web/packages/tm/index.html) package, though
powerful, was frustrating, partly due to its unusual data format, which was
tricky to traverse.  But these new packages operate on ordinary data frames,
using nesting to great effect.

The outcome is that n-grams can be created incredibly quickly, easily avoiding
sentence boundaries.  The code is as simple as this:

```{r harry-potter-sentiment-ngramming, eval = FALSE, echo = TRUE}
readRDS(here("data", "books-raw.Rds")) %>%
  # One row per paragraph. Two columns: title and text
  # First, break into sentences so that ngrams don't cross sentence boundaries
  group_by(title) %>%
  summarise(sentence = list(unlist(map(text, tokenize_sentences)))) %>%
  unnest %>%
  # Then create 4-grams
  group_by(title) %>%
  summarise(fourgram = list(unlist(tokenize_ngrams(sentence, n = 4)))) %>%
  unnest
```

## Most-common phrases in *Harry Potter*

Did you hear that the most common phrase in *Harry Potter* is "turned on his
heel"?  I can finally bust that myth.  It does appear quite often -- 12 times at
most, if you include 'turning' and 'her', but the most-common four-word phrase,
by *miles*, is "Harry, Ron and Hermione".  Big surprise.

```{r harry-potter-sentiment-fourgrams, echo = FALSE}
readRDS(here("data", "ngrams.Rds")) %>%
  filter(words == 4) %>%
  arrange(rank) %>%
  slice(1:30) %>%
  arrange(desc(rank)) %>%
  mutate(ngram = factor(ngram, levels = ngram)) %>%
  ggplot(aes(ngram, n)) +
  coord_flip() +
  geom_text(aes(label = ngram), hjust = 0) +
  scale_y_continuous(limits = c(0, 310), expand = c(0, 0)) +
  xlab("") +
  theme_minimal() +
  ylab("Occurrences (left-justified)") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank()) +
  ggtitle("Top 30 Harry Potter 4-grams")
```

A few of these are predictable nouns (Defence Against the Dark Arts, the
Ministry of Magic, the Room of Requirement).  He Who Must Not Be Named makes it
into the top 40.  There are a bunch of phrases that describe where things are
(at the end of, etc.).  But the most intriguing phrase is "said Hermione in a"
-- why is Hermione singled out by that construction?

There's a [shiny
app](https://duncan-garmonsway.shinyapps.io/harry-potter/#n-grams) to explore
lots more n-grams, from 2-grams to 10-grams.

[Slate](http://www.slate.com/articles/arts/culturebox/2013/11/hunger_games_catching_fire_a_textual_analysis_of_suzanne_collins_novels.html?wpisrc=burger_bar)
did a similar analysis, though they looked at the most-common *sentences*,
comparing *Harry Potter* with *The Hunger Games* and the *Twilight* series.

They seem to have edited their list somewhat, since "He waited" appears only
three times, and "Something he didn't have last time" only twice, while I find
that "Harry nodded" tops my list (of complete sentences) with 14 occurences,
one more than Slate's top sentence, "Nothing happened."

Here are my top 30, many of which are not complete sentences.

```{r harry-potter-sentiment-sentences, echo = FALSE}
sentences <-
  readRDS(here("data", "sentences.Rds")) %>%
  count(sentence) %>%
  arrange(desc(n)) %>%
  mutate(rank = min_rank(-n))

sentences %>%
  slice(1:30) %>%
  arrange(desc(rank)) %>%
  mutate(sentence = factor(sentence, levels = sentence)) %>%
  ggplot(aes(sentence, n)) +
  coord_flip() +
  geom_text(aes(label = sentence), hjust = 0) +
  scale_y_continuous(limits = c(0, 61), expand = c(0, 0)) +
  xlab("") +
  ylab("Occurrences (left-justified)") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank()) +
  ggtitle("Top 30 Harry Potter sentences")
```

Part of the difficulty is that written English speech isn't unambiguously
punctuated.  This has bugged me since primary school.  See what happens here.

```{r harry-potter-sentiment-ambiguous, echo = TRUE}
tokenize_sentences(c("'Are you going?' Harry asked.",
                     "Ron asked, 'Are you going?' Harry shrugged.",
                     "'You should go,' Harry said",
                     "'Go now.' Harry went."))
```

Were I king, I'd decree the following unambiguous style.

```{r harry-potter-sentiment-decree, echo = TRUE}
tokenize_sentences(c("Ron asked, 'Are you going?'. Harry shrugged.",
                     "'You should go.', Harry said"))
```

## Most-important characters

If importance is proportional to mentions of first names, then Hermione and Ron
are not as equal as you might expect.

```{r harry-potter-sentiment-characters, echo = FALSE, message = FALSE}
books <- readRDS(here("data", "books.Rds"))

booksplot <-
  books %>%
  ungroup %>%
  # Count occurrences of name
  arrange(title, chapter, linenumber, wordnumber) %>%
  mutate(harry = cumsum(stri_detect(word, regex = "harry")),
         ron = cumsum(stri_detect(word, regex = "ron")),
         hermione = cumsum(stri_detect(word, regex = "hermione"))) %>%
  # Take only the end-of-chapter numbers (fewer data points for slow ggplot2)
  group_by(title, chapter, linenumber) %>%
  summarise_at(vars(harry, ron, hermione), last) %>%
  ungroup %>%
  # Plot
  mutate(linenumber = row_number()) %>%
  gather(character, mentions, harry, ron, hermione) %>%
  mutate(character = factor(character,
                            levels = c("harry", "ron", "hermione"),
                            labels = c("Harry ", "Ron ", "Hermione "))) %>%
  ggplot(aes(linenumber, mentions, colour = character)) +
  geom_line() +
  scale_colour_discrete("") +
  xlim(0, 40000) +
  xlab("Progress through the series") +
  ylab("Number of mentions of name (the exact name in the legend)") +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank())
booksplot %>% direct.label(method = "last.polygons")
```

## Sentiment-driven story arcs

Ever since I read [Julia Silge](http://juliasilge.com/blog/)'s amazing story-arc
[sentiment](http://juliasilge.com/blog/Life-Changing-Magic/)
[analyses](http://juliasilge.com/blog/Life-Changing-Magic/), I wanted to apply
the method to the *Harry Potter* books.

There's a [shiny
app](https://duncan-garmonsway.shinyapps.io/harry-potter/#sentiment) to explore
this interactively, but here is a still for the blog.

```{r harry-potter-sentiment-arcs, echo = FALSE}
arcs <- readRDS(here("data", "dashboard", "arcs.Rds"))

ranges <-
  arcs %>%
  filter(low_pass_size >= 3, low_pass_size <= 10) %>%
  group_by(title, row) %>%
  summarise(min = min(transformed), max = max(transformed))

main <-
  ggplot() +
  ylim(-3.5, 3.5) +
  facet_wrap(~title, ncol = 2, dir = "v") +
  theme_minimal() +
  ylab("Transformed Sentiment Value") +
  labs(title = "Sentiment in Harry Potter") +
  scale_x_discrete(expand=c(0,0)) +
  theme(plot.title = element_text(size = 20)) +
  theme(strip.text = element_text(hjust = 0)) +
  theme(strip.text = element_text(face = "italic")) +
  theme(strip.text = element_text(size = 14)) +
  theme(axis.text.y = element_text(margin = margin(r = -10))) +
  theme(legend.position = "none") +
  theme(axis.title.y = element_text(size = 14)) +
  theme(axis.title.x = element_blank()) +
  theme(axis.ticks.x = element_blank()) +
  theme(axis.text.x = element_blank())

hline <-
  geom_hline(yintercept = 0, alpha = .3)

ribbon <-
  geom_ribbon(aes(row, ymin = min, ymax = max),
              fill = "grey85",
              data = ranges)

arcline <-
  geom_line(aes(row, transformed),
            colour = "#2780E3",
            size = 1,
            data = arcs %>% filter(low_pass_size == 3))

arcarea <-
  geom_area(aes(row, transformed),
            fill = "#2780E3",
            alpha = .5,
            data = arcs %>% filter(low_pass_size == 3))

main_arcarea <- main + arcarea
main_ribbon <- main + ribbon + hline
main_ribbon_arcline <- main + ribbon + arcline + hline
```

```{r harry-potter-sentiment-arcarea, echo = FALSE}
main_arcarea
```

If there's anything to interpret here, then it's that the first three books
play the game "fortunately, unfortunately", while the later books are a little
different, especially *Order of the Phoenix*, which is the grumpy one.

Perhaps the Fourier transform is too sensitive to a magic number that I call the
'wiggliness' parameter.  To see how sensitive, I calculated the arcs for
'wiggliness' values from 3 to 10, and described the range of the arcs with a
ribbon -- a little like the standard-error-ribbon on a `geom_smooth`.  I think
the ribbons show a more reliable story arc, and reveal that narrow wobbles, of
the order of a chapter or so, are probably misleading.

```{r harry-potter-sentiment-ribbon, echo = FALSE}
main_ribbon
```

And to see where the 'wiggliness = 3' arc lies in the range, I superimpose it as a line.

```{r harry-potter-sentiment-ribbon-arcline, echo = FALSE}
main_ribbon_arcline
```

The code is, as always, on [GitHub](https://github.com/nacnudus/crossprod), but
you need to supply your own copies of the books.

