---
title: "{ggstraw}: A custom ggplot2 geom for deviations"
description: |
  ggstraw::geom_straw() is a custom ggplot2 geom for showing deviations from
  expected values, e.g. actual vs scheduled departure times.
author: Duncan Garmonsway
date: January 25, 2019
output:
  radix::radix_article:
    self_contained: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(tidyverse)
```

Commuters on public transport are acutely aware of deviations from expected
values.  Plane departures, for example.  How many planes depart on time?  And
how can you make that statistic vivid?

One way is to visualise each plane's deviation from the schedule.  Here is a
graph of three days when 83% of planes departed on time.  It uses
[`ggstraw::geom_straw()`](https://github.com/nacnudus/ggstraw).

```{r flights}
library(tidyverse)
library(hms)
library(ggplot2)
library(ggstraw)

ggstraw::flights %>%
  ggplot(aes(sched_dep_time, carrier, xend = dep_time)) +
  geom_straw() +
  xlab("") +
  ylab("Carrier") +
  ggtitle("Departure times from New York to Boston", sub = "Scheduled vs actual") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

That graph shows every plane departure, not just a summary statistics.  Each
departure is a near-vertical line that slopes from the scheduled departure time
at its base on the x-axis to the actual departure time at its tip above the
x-axis.  The neat vertical lines of the carrier "US" are comfortable to read,
whereas  the awkwardly sloped lines of the carriers "UA" and "B6" are
uncomfortable to read and demand attention.  This reflects the fact that the
uncomfortable sloping lines indicate late flights, deviations from the expected
schedule.

Early flights would also be uncomfortable to read and demand attention.  I think
that is also desirable, particularly in forms of transport that abandon
passengers by departing early, such as trains and buses.

Because every plain departure is shown, they are easy to relate to. You can see
what time of day planes were late, to help you decide which time of day to fly.
You can also get a feel for how late the planes were.

Thanks to [ggplot2's magical extension system](https://ggplot2.tidyverse.org/articles/extending-ggplot2.html) you get things like animation for free.

```{r animation}
library(gganimate)
library(lubridate)

animation <-
  ggstraw::flights %>%
  mutate(day = floor_date(sched_dep_time, "day"),
         sched_dep_time = as.hms(sched_dep_time),
         dep_time = as.hms(dep_time)) %>%
  ggplot(aes(sched_dep_time, carrier, xend = dep_time)) +
  geom_straw() +
  transition_states(day, transition_length = 0.5, state_length = 0.5) +
  enter_fade() +
  exit_shrink() +
  ease_aes('sine-in-out') +
  xlab("") +
  ylab("Carrier") +
  ggtitle("Departure times from New York to Boston",
          sub = "Scheduled vs actual\n{previous_state}") +
  theme_minimal() +
  theme(panel.grid = element_blank())

animate(animation, nframes = 25, fps = 5)
```

Another use is to show when 'fast' trains overtake 'all-stations' trains.  Here
we put the scheduled start time at the bottom of each straw, and the scheduled
arrival time at the top.  Crossed straws draw the eye to the overtakes. Here is
a plot of part of the London Underground Metropolitan Line timetable.

```{r timetable}
ggstraw::timetable %>%
  ggplot(aes(x = departure, xend = arrival,
             y = day)) +
  geom_straw() +
  xlab("") +
  ylab("") +
  ggtitle("Metropolitan timetable from Harrow-on-the-Hill to Moor Park",
  subtitle = "Scheduled departure time (bottom) and scheduled arrival time (top)") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

Look again and you might notice a gap in arrivals at about 19:15 on weeknights.
That one often catches me out -- for some reason several trains in a row take a
different branch, to Uxbridge, while commuters travelling towards Moor Park,
Amersham, Chesham and Watford pass the time writing blog posts about R.
