---
title: "Creating nests without tidyr"
description: |
  "How to construct nested data frames from scratch, without tidyr"
author: "Duncan Garmonsway"
date: June 22, 2016
output:
  radix::radix_article:
    self_contained: false
resources:
  exclude:
    data
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "##", collapse = TRUE, cache = TRUE)
```

Unless you begin with an unnested data frame, creating a nested data frame needs
a little trick.  Here it is.

# Nested data frames

The `tidyr` package has a handy function for *nesting* data frames.  Hadley
Wickham describes it [thus](https://blog.rstudio.org/2016/02/02/tidyr-0-4-0/):

> In a grouped data frame, you have one row per observation, and additional
> metadata define the groups. In a nested data frame, you have one row per
> group, and the individual observations are stored in a column that is a list
> of data frames. This is a useful structure when you have lists of other
> objects (like models) with one element per group.

Here's a small example:

```{r nests-iris, echo = TRUE}
library(dplyr)
library(tidyr)

iris_nested <-
  iris %>%
  group_by(Species) %>%
  sample_n(2) %>%
  nest
iris_nested
iris_nested %>% str
```

Interestingly, the nested column isn't a vector like ordinary columns; it's a
list.  Actually lists are just one kind of vector -- the non-atomic kind
(composed of parts, i.e vectors and other lists), whereas
integer/character/etc. vectors are the atomic kind (not composed of parts).
This is nicely explained in [Advanced
R](http://adv-r.had.co.nz/Data-structures.html#vectors) by Hadley Wickham.

```{r nest-vector-list, echo = TRUE}
is.atomic(vector(mode = "character", length = 2))
is.atomic(vector(mode = "list", length = 2))
```

# Please say it's a data frame

Data frames, which are a list of vectors, handle list-type columns perfectly
well, but data-frame-construction functions don't. So when I tried to create
one from scratch (rather than by nesting an existing data frame as above), I
lost a lot of time mucking about with `data.frame()` and the like.

```{r nests-mucking-about, echo = TRUE, error = TRUE}
data.frame(X1 = 1:2, X2 = list(iris, mtcars))
as.data.frame(list(X1 = 1:2, X2 = list(iris, mtcars)))
```

# It's a data frame because I say so

The answer is to simply tell R that the data structure is a data frame by
setting its class and giving it a "row.names" attribute.

```{r nests-construction, echo = TRUE}
x <- list(X1 = 1:2, X2 = list(iris[1:2, 1:2], iris[3:5, 1:4]))
structure(x, class = c("tbl_df", "data.frame"), row.names = 1:2)
```

# Invading the nest

Accessing the nested column by the usual subsetting
operators, `$`, `[` and `[[`, is a little clumsy.

```{r nests-subsetting, echo = TRUE}
x$X2 # Returns the list of data frames
x$X2[2] # Returns the second data frame, wrapped in a list

x$X2[[2]] # Returns the second data frame -- probably what you want
x[["X2"]][[2]] # Returns the second data frame -- also probably what you want
```

***[Edit: The code below no longer works in the latest version of R.]***

```{r nests-subsetting2, echo = TRUE, error = TRUE}
x[2, "X2"] # Returns the second data frame wrapped in another data frame
x[2, "X2", drop = TRUE] # Same -- ignores "drop"
x[2, "X2"][1, "X2"][1, "X2"] # The `[` goes around in circles
```

