---
title: "Add nowt() to your tidy pipelines"
description: |
  A simple function for your .Rprofile to make interactive coding with pipelines
  less of a faff.
author: Duncan Garmonsway
date: January 31, 2019
output:
  radix::radix_article:
    self_contained: false
preview: ./nowt-so-queer-as-folk.jpg
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Do you write perfect tidyverse pipelines first time?  Then like Cédric Scherer
[\@CedScher](https://twitter.com/CedScherer/status/1090608766842798080) and Wei
Yang Tham [\@wytham88](https://twitter.com/wytham88/status/1012085073368272903)
you probably find yourself faffing with trailing `%>%` when removing or
commenting the final operation in a pipeline.

One way to get out of it is to
append `identity()` or a rather cryptic `{.}` to the end of every pipeline.

```{r}
# Experimental code
iris %>%
  group_by(Species) %>%
  summarise(Mean.Sepal.Length = mean(Sepal.Length)) %>%
  filter(Mean.Sepal.Length > 6)
```

```{r, error = TRUE}
# Comment out the final step
iris %>%
  group_by(Species) %>%
  summarise(Mean.Sepal.Length = mean(Sepal.Length)) %>%
  # filter(Mean.Sepal.Length > 6)
```

```{r}
# Oops, forgot to deal with the trailing %>%
iris %>%
  group_by(Species) %>%
  summarise(Mean.Sepal.Length = mean(Sepal.Length))
```

```{r}
# Could have been avoided with identity() or {.}
iris %>%
  group_by(Species) %>%
  summarise(Mean.Sepal.Length = mean(Sepal.Length)) %>%
  # filter(Mean.Sepal.Length > 6) %>%
  identity()

# Adapted from @CedScherer's tweet
```

Unfortunately `identity()` and `{.}` don't work in ggplot2 piplines, so you have
to use `NULL` instead.

```{r}
ggplot(iris, aes(Sepal.Length, Sepal.Width)) +
  geom_point() +
  # scale_y_continuous(position = "right") +
  NULL
```

We can do better.  How about a single function for both situations, with documentation and a helpful name that means "nothing" in Northern English: `nowt()`.

```{r}
nowt <- function(...) {
  if (nargs()) {
    return(identity(...))
  }
  NULL
}

iris %>%
  group_by(Species) %>%
  summarise(Mean.Sepal.Length = mean(Sepal.Length)) %>%
  # filter(Mean.Sepal.Length > 6) %>%
  nowt()

ggplot(iris, aes(Sepal.Length, Sepal.Width)) +
  geom_point() +
  # scale_y_continuous(position = "right") +
  nowt()
```

Okay, so I didn't document it because it's in the .Rprofile rather than a
packge.  But if this idea were picked up by the tidyverse then it _could_ have
documentation, unlike `NULL` or `{.}`.
