---
title: "More Harry Potter story arcs (they get darker)"
description: |
  Reproducing *Harry Potter* sentiment arcs by some real wizards
author: Duncan Garmonsway
date: July 20, 2016
preview: "harry-potter-arcs-wizards.jpg"
output:
  radix::radix_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, cache = TRUE)
```

```{r harry-potter-arcs-packages, echo = FALSE}
library(purrr)
library(dplyr)
library(tidyr)
library(stringi)
library(ggplot2)
library(scales)
library(qdapDictionaries) # labMT dictionary
library(tidytext)
library(tokenizers)
library(here)
```

In my [previous
post](https://nacnudus.github.io/duncangarmonsway/posts/2016-07-13-harry-potter-sentiment/),
I applied [Julia Silge](http://juliasilge.com/blog/)'s story-arc
[sentiment](http://juliasilge.com/blog/Life-Changing-Magic/)
[analysis](http://juliasilge.com/blog/Life-Changing-Magic/) method to the *Harry
Potter* books. Twitter's response was a tweet about some very similar and
fascinating research, on [arXiv](https://arxiv.org/abs/1606.07772).  Here's
their beautiful graph.

![*Harry Potter and the Deathly Hallows* sentiment arc, from Reagan et al.
https://arxiv.org/pdf/1606.07772v2.pdf](harry-potter-arcs-wizards.jpg)

This post reproduces their method, using R.

### Method

Reagan et al.'s method differs in two ways. They use the sentiment dictionary
LabMT, a product of their own research group,
[Hedonometer](http://hedonometer.org/index.html), which is available in R in the
[qdapDictionaries](https://github.com/trinker/qdapDictionaries) package.  Then
they smooth the scores by averaging inside a moving window.  The moving-window
method produces curves that make a lot more sense.

You can see their curves for all the Harry Potter books on their interactive
[website](http://hedonometer.org/books/v1/?book=Harry%20Potter%20and%20the%20Sorcerer%27s%20Stone),
which I highly recommend.

But I wanted to reproduce the method in R, for fun.  The main snag was handling
the moving window, which is one of the few under-developed parts of R munging at
the moment.  There's [RcppRoll](https://github.com/kevinushey/RcppRoll), which
implements a few obvious functions but no-longer allows custom functions, and
[zoo](https://cran.r-project.org/web/packages/zoo/index.html), which does allow
custom functions, but seems to want the input to be a timeseries -- fair enough,
it is a timeseries library.

So I did something hacky with base R and
[purrr](https://github.com/hadley/purrr).  Starting with two vectors, one for
the first and the other for the last row number of each window.  Putting those
into a list, I then **transpose**d them, which you really have to see to
understand what it's doing -- basically it paired up each entry of each vector
into a vectors of length 2, each one an element of one overall list.  Then I
could use **lapply** to iterate over each pair, generating a sequence of
rownumbers from the first to the last, and finally **bind_rows** combined them
into a large dataframe of all rows in all windows.  I recommend running the code
below to see how this works.  Obviously the method doesn't scale too well, since
it loads all the windows into memory at once.

```{r, echo = TRUE, eval = FALSE}
# Set up the moving-window-average parameters
N <- nrow(book) # Number of words in the book
k <- 10000 # Number of words in each sample
l <- 200 # Number of points in the time series
overlap <- floor((N - k)/(l - 1)) # Number of words by which the window slides
overlap <- floor((N - k - 1)/l) # Function in the paper -- I think it's wrong, but it makes little difference.

# first and last words of each chunk
uppers <- c(seq(k, by = overlap, length.out = l), N)
lowers <- c(seq(1, by = overlap, length.out = l), N - k)

arc <-
  list(lowers, uppers) %>%
  transpose %>%
  at_depth(1, unlist) %>%
  lapply(function(x) {slice(book, x[1]:x[2]) %>% mutate(length = n())}) %>%
  bind_rows(.id = "window") %>%
  etc.
```

One thing on my graphs that Hedonometer doesn't provide is the chapter titles,
so you can be reminded what is going on at each turn in the plot.

Enjoy the plots below, otherwise that's it for this post.  The code is, as
always, on
[GitHub](https://github.com/nacnudus/duncangarmonsway/blob/master/_posts/2016-07-20-harry-potter-arcs/2016-07-20-harry-potter-arcs.Rmd),
but you need to supply your own copies of the books.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Titles as factor in series order
titles <- c("1. Philosopher's Stone",
            "2. Chamber of Secrets",
            "3. Prisoner of Azkaban",
            "4. Goblet of Fire",
            "5. Order of the Phoenix",
            "6. Half-Blood Prince",
            "7. Deathly Hallows")

# Filenames (alphanumerically in series order)
files <-
  list.files(here("..", "2016-07-13-harry-potter-sentiment", "data", "txt"), full.names = TRUE) %>%
  normalizePath

# Load the texts (one paragraph per row)
books <-
  tibble(title = titles, file = files) %>%
  mutate(text = map(file, readLines)) %>%
  dplyr::select(-`file`) %>%
  unnest %>%
  filter(text != "")

# Mark chapters (not really necessary for this script)
# Start-of-book markers: ^— CHAPTER ONE —$ but with different kinds of -.
# End-of-book markers: ^Titles available.*
chapters <- c("ONE", "TWO", "THREE", "FOUR", "FIVE", "SIX", "SEVEN", "EIGHT",
              "NINE", "TEN", "ELEVEN", "TWELVE", "THIRTEEN", "FOURTEEN",
              "FIFTEEN", "SIXTEEN", "SEVENTEEN", "EIGHTEEN", "NINETEEN",
              "TWENTY", "TWENTY-ONE", "TWENTY-TWO", "TWENTY-THREE",
              "TWENTY-FOUR", "TWENTY-FIVE", "TWENTY-SIX", "TWENTY-SEVEN",
              "TWENTY-EIGHT", "TWENTY-NINE", "THIRTY", "THIRTY-ONE",
              "THIRTY-TWO", "THIRTY-THREE", "THIRTY-FOUR", "THIRTY-FIVE",
              "THIRTY-SIX", "THIRTY-SEVEN", "THIRTY-EIGHT",
              "Nineteen Years Later")
books <-
  books %>%
  group_by(title) %>%
  mutate(chapter = stri_extract(text, regex = "^[—–] CHAPTER .*|^Titles available.*|Nineteen Years Later"),
         chapter_title = if_else(is.na(chapter), as.character(NA), lead(text, 1)),
         chapter_title = if_else(chapter %s==% "Nineteen Years Later", chapter, chapter_title)) %>%
  fill(chapter, chapter_title) %>%
  filter(!is.na(chapter)) %>%
  filter(cumsum(stri_detect(chapter, regex = "^Titles available.*")) == 0) %>%
  # Convert chapter headings to integers
  mutate(chapter = stri_extract(chapter, regex = "[a-zA-Z -]+"),
         chapter = stri_replace_first_fixed(chapter, "CHAPTER", ""),
         chapter = stri_trim(chapter)) %>%
  mutate(chapter = as.integer(factor(chapter, levels = chapters))) %>%
  mutate(linenumber = row_number()) %>%
  filter(linenumber >= 3) %>%
  ungroup()

# Convert non-ascii characters for tokenizing and ’ to ' to match stopwords
books <-
  books %>%
  mutate(text = stri_replace_all(text, replacement = "'", regex = "[‘’]"),
         text = stri_replace_all(text, replacement = "-", regex = "[–?]"))

# Prepare for sentiment analysis ===============================================

# Convert to one token per row
books <-
  books %>%
  unnest_tokens(word, text) %>%
  group_by(title, chapter, chapter_title, linenumber) %>%
  mutate(wordnumber = row_number()) %>%
  ungroup

# Sentiment arcs ---------------------------------------------------------------

arc <- function(book) {
  book <-
    book %>%
    ungroup %>%
    arrange(chapter, chapter_title, linenumber, wordnumber) %>%
    mutate(wordnumber = row_number()) %>%
    dplyr::select(chapter, chapter_title, wordnumber, word)

  chapter_titles <-
    book %>%
    group_by(chapter, chapter_title) %>%
    summarise(words = n()) %>%
    ungroup %>%
    mutate(percent = cumsum(words/sum(words)),
           centre = lag(percent, default = 0) + ((percent - lag(percent, default = 0)) / 2))

  N <- nrow(book) # Number of words in the book
  k <- 10000 # Number of words in each sample
  l <- 200 # Number of points in the time series
  overlap <- floor((N - k)/(l - 1)) # Number of words by which the window slides
  overlap <- floor((N - k - 1)/l) # Function in the paper -- I think it's wrong, but it makes little difference.

  # first and last words of each chunk
  uppers <- c(seq(k, by = overlap, length.out = l), N)
  lowers <- c(seq(1, by = overlap, length.out = l), N - k)

  arc <-
    list(lowers, uppers) %>%
    transpose %>%
    at_depth(1, unlist) %>%
    lapply(function(x) {slice(book, x[1]:x[2]) %>% mutate(length = n())}) %>%
    bind_rows(.id = "window") %>%
    inner_join(labMT, by = "word") %>%
    dplyr::select(window, length, chapter, word, h_avg = happiness_average) %>%
    group_by(window, length, word) %>%
    summarise(chapter = max(chapter),
              h_avg = first(h_avg),
              f = n()) %>%
    mutate(p = f/sum(f)) %>%
    group_by(window) %>%
    summarise(length = first(length),
              chapter = first(chapter),
              h_avg = sum(h_avg * p, na.rm = TRUE)) %>%
    mutate(window = as.integer(window)) %>%
    arrange(window)
}

arcs <-
  books %>%
  group_by(title) %>%
  do(arc(.))

chapter_titles <-
  books %>%
  group_by(title, chapter, chapter_title) %>%
  summarise(words = n()) %>%
  group_by(title) %>%
  mutate(percent = cumsum(words/sum(words)),
         centre = lag(percent, default = 0) + ((percent - lag(percent, default = 0)) / 2))

plotbook <- function(book) {
  chapter_titles <-
    books %>%
    filter(title == book$title[1]) %>%
    group_by(chapter, chapter_title) %>%
    summarise(words = n()) %>%
    ungroup %>%
    mutate(percent = cumsum(words/sum(words)),
           centre = lag(percent, default = 0) + ((percent - lag(percent, default = 0)) / 2))
  bookplot <-
    book %>%
    mutate(x = cumsum(length) / sum(length)) %>%
    ggplot(aes(x, h_avg)) +
    geom_vline(aes(xintercept = percent),
               data = chapter_titles,
               colour = "grey50") +
    geom_text(aes(x = centre, label = chapter_title),
              y = min(arcs$h_avg),
              data = chapter_titles,
              angle = 90,
              hjust = 0,
              colour = "grey50",
              size = 3) +
    geom_hline(yintercept = mean(arcs$h_avg), colour = "blue", linetype = 2, size = 1) +
    geom_line() +
    ylab("Happiness") +
    xlab("") +
    scale_x_continuous(labels = percent, expand = c(0, .01)) +
    facet_wrap(~title, dir = "v")
  print(bookplot)
  return(data.frame())
}

arcs %>%
  group_by(title) %>%
  do(plotbook(.))
```

