---
title: "Simmer vs SimPy: The Bank, Part I"
description: |
  Porting a SimPy tutorial to the R package Simmer
author: Duncan Garmonsway
date: May 11, 2016
output:
  distill::distill_article:
    self_contained: false
resources:
  exclude:
    data
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "##", collapse = TRUE, cache = TRUE)
```

***A maintained version of this post is available with
[simmer](http://r-simmer.org/) as a
[vignette](http://r-simmer.org/articles/D-bank-1.html).  Simmer has undergone
heavy development since this post was written.***

## Simmer vs SimPy

Which package would be easier for teaching queueing theory?  Python 2.7's
[`SimPy`](https://pythonhosted.org/SimPy/SimPy_Overview.html), designed for
(as far as I can tell) lecturing, by Tony Vigneau at my alma mater, Vic Uni
Wellington NZ, or [`simmer`](http://r-simmer.org/), designed by Bart Smeets and
Iñaki Ucar to (as far as I can tell) actually use?

The `simmer` package is a relatively new R package for discrete event simulation
(DES).  It's an exciting development, because there isn't a lot of open-source
DES software.  `SimPy` seems to be the only serious competitor for teaching
DES and queueing theory.

This post does three things:

* Ports the code of the main `SimPy` tutorial ['The
  Bank'](https://pythonhosted.org/SimPy/Tutorials/TheBank.html) to `simmer`.
* Opines that `simmer` would be easier to teach as part of a queueing theory
  course.
* Pursues a random red herring.

## Why not SimPy 3?

I use `SimPy` 2 (for Python 2), because it is the last version developed by the
original author, because it was the version I was taught, only last year, and
because, in one crucial respect (monitoring), it's [much easier to
use](https://simpy.readthedocs.io/en/latest/topical_guides/monitoring.html).

## The Bank

'The Bank' is a tutorial that develops DES concepts and techniques by simulating
the paths of customers at a bank.  The arrivals (customers) queue for a server
(counter), are served, and exit.

## Complete example

The actual ported code is available [on
GitHub](https://github.com/nacnudus/simmer-bank), and I only give simple
examples in this post.  The first example is complete.

First, `SimPy`:

```{r, engine = "python", engine.path = "/usr/bin/python2", echo = TRUE}
""" bank01: The single non-random Customer """
from SimPy.Simulation import *

## Model components -----------------------------

class Customer(Process):
    """ Customer arrives, looks around and leaves """

    def visit(self,timeInBank):
        print now(),self.name," Here I am"
        yield hold,self,timeInBank
        print now(),self.name," I must leave"

## Experiment data ------------------------------

maxTime = 100.0     # minutes
timeInBank = 10.0   # minutes

## Model/Experiment ------------------------------

initialize()
c = Customer(name="Klaus")
activate(c,c.visit(timeInBank),at=5.0)
simulate(until=maxTime)
```

Next, `simmer`:

```{r, echo = TRUE}
# bank01: The single non-random customer
suppressMessages(library(simmer))

## Experiment data ------------------------------

maxTime <- 100     # minutes
timeInBank <- 10   # minutes

## Model components -----------------------------

customer <-
  create_trajectory("Customer's path") %>%
  timeout(function() {timeInBank})

## Model/Experiment ------------------------------

bank <- simmer("bank")
bank %>% add_generator("Customer", customer, at(5))
bank %>% run(until = maxTime)
bank %>% get_mon_arrivals
```

Already there are several differences that might make teaching queueing theory
with `simmer` easier than with `SimPy`:

* The difference between `from X import Y` and `Import X` isn't relevant.
* Whitespace doesn't matter.
* The difference between integer and floating-point types doesn't matter here.
* Arguments don't have to be defined.
* References don't have to be passed.
* `self` is irrelevant.
* `timeout` is more intuitive than `yield` (`yield` describes how the class
  behaves in the implementation of the DES, as it yields control back to the
  clock, whereas `timeout` describes what the function does in the mind of the
  modeller).

But there is one point that could be tricky, and that soon becomes important:

* `timeout` and `add_generator` both expect *functions*, rather than vectors, to
  control (inter-)arrival time and timeout duration.  It would be nice to have
  syntactic sugar to handle vectors.  The reason for the functions is that, when
  a model is run indefinitely, a function can continue generating new arrival
  times and timeout durations, whereas a vector will soon be exhausted.

## Example fragments

Implementing the rest of the examples brought up a few other interesting points.

### Generate more than one arrival

In the `SimPy` examples, to generate `n > 1` arrivals, the `activate` code to
generate them moves inside the `Source` class.  To explain why requires a quite
a lot of understanding/intuition of object-oriented programming that isn't
relevant to learning about queuing theory.  `Simmer` doesn't present this
difficulty.

### Limit the number of arrivals

Arrivals with random inter-arrival times would be generated indefinitely by
`bank %>% add_generator("Customer", customer, function() {runif(1)})`.  To limit
this to `n = 10` arrivals, you might try `times <- runif(10); bank %>%
add_generator("Customer", customer, times)`, but it doesn't work, because
`add_generator` expects a function that *will* supply inter-arrival times, not a
vector that *does* supply them.

`Simmer` provides a handy function, `at()`, to
convert a vector to function, so you could do `add_generator("Customer", customer,
at(runif(10)))`, except that this still doesn't work.  That's because `at()`
is designed to convert *arrival* times into *inter-arrival* times, but the
`runif` function is being used to provide *inter-arrival* times in the first
place.  The final fix is to do `add_generator("Customer", customer, at(c(0,
cumsum(runif(10)))))`.

### Joining the shortest queue

This is a pain in both `SimPy` and `simmer`.  The `SimPy` example creates a
method to return the length of each queue, and then the following code iterates
through the results until a queue is chosen:

```python3
# Select the shortest queue
for i in range(Nc):
    if Qlength[i] == 0 or Qlength[i] == min(Qlength):
        choice = i  # the chosen queue number
        break

# Join the queue
yield request,self,counters[choice]
```

In `simmer`, this is done by branching:

```r
customer <-
  create_trajectory("Customer's path") %>%
    branch(function() {
             # Select the shortest queue
             which.min(c(bank %>% get_server_count("counter1") +
                           bank %>% get_queue_count("counter1"),
                         bank %>% get_server_count("counter2") +
                           bank %>% get_queue_count("counter2")))
         },
         merge = rep(TRUE, 2),
         # Join the first queue, if it was chosen
         create_trajectory("branch1") %>%
           seize("counter1") %>%
           timeout(function() {rexp(1, 1/timeInBank)}) %>%
           release("counter1"),
         # Otherwise join the second queue, if it was chosen
         create_trajectory("branch2") %>%
           seize("counter2") %>%
           timeout(function() {rexp(1, 1/timeInBank)}) %>%
           release("counter2"))
```

I mucked about for a while trying to avoid branching by using attributes to name
the server at `seize` time. I won't explain attributes here because they're
covered in the excellent `simmer` [vignettes](), but basically the following
code doesn't work because attributes are only available to certain arguments,
the `resource` argument not among them, only `amount` and perhaps
`priority` and `preemptible`.

```r
# This doesn't work:
customer <-
  create_trajectory("Customer's path") %>%
    # Attributes can be set, to choose the queue
    set_attribute("counter",
                  function() {
                    which.min(c(bank %>% get_server_count("counter1") +
                                  bank %>% get_queue_count("counter1"),
                                bank %>% get_server_count("counter2") +
                                  bank %>% get_queue_count("counter2")))}) %>%
    # But they aren't available in the `resource` argument of `seize` for naming
    # the server, so this doesn't work.
    seize(function(attrs) {paste0("counter", attrs["counter"])}) %>%
    timeout(function() {rexp(1, 1/timeInBank)}) %>%
    release(function(attrs) {paste0("counter", attrs["counter"])})
```

### Monitoring

`Simmer` has a killer feature: everything is monitored automatically, and
reported in handy data frames.  This works especially well when doing many
replications.

But it isn't obvious how to do the equivalent of, in Python, injecting `print`
or `cat` commands to describe the state of particular arrivals and servers.
Presumably something could be done in the functions passed to `dist` arguments.
In this sence, `simmer` is more declarative; like a story book, where the text
describes the characters, but the characters don't really exist.  `Simmer`
describes arrivals and servers, but they don't really exist, and can't be
directly interacted with.

## Random red herring

Python 2.7, R and MATLAB all use the Mersenne-Twister algorithm by default.  But
none of them matches.  The `numpy` Python package [*does* match
MATLAB](http://www.walkingrandomly.com/?p=5479) (except for seed = 0), but not
R.

Two potential solutions are:

* Generate any old random numbers, write them to disk, and read them into both
  Python and R.
* Use `rpy2` to use R's random number generator from within Python.

I used `rpy2`, but it wasn't long before I encountered a more serious problem.
When random draws are conducted in more than one part of the code, the
programmer can't control the order of the draws.  That's up to `SimPy` and
`simmer`.  At that point, I gave up.

